### 2011-10-27 contents.tex
Initial paper information and outline.

\section{Introduction}

\section{Background}

\subsection{Bug Tracking Systems}
Bug workflow. Bugzilla. \cite{Bird2009}

\subsection{The Eclipse Project}
History. \\
Recommendations about Bugzilla usage. Four-eyes principle.

\subsection{The Netbeans Project}
History. \\
Recommendations for developers.

\section{Experimental Setup}

\begin{verbatim}
Data extraction
Data processing: 
- filtering (bugs up to december 20xx)
- classification script
Sampling:
- random sampling (for coding)
- entire data set
Analysis:
- grounded theory (open coding)
- latent Dirichlet allocation
- descriptive statistics
\end{verbatim}

\section{Results}

\section{Conclusion}

\section*{Acknowledgment}

The first author is supported by FAPESB under grant BOL0119/2010.

### 2011-10-27 contents.tex
Adding snippets from tumblr and the technical report.

@@
-\section{Introduction}
+\begin{section}{Introduction}
 
-\section{Background}
+Software development is a complex activity that demands the coordination of many people, such as programmers, designers, and managers.
+Software development activities should be coordinated in order to deliver products with high quality and on schedule.
+Many practices and processes have been proposed in order to aid the software development, targeting at high quality, low cost, fast delivery, or a combination of these.
 
-\subsection{Bug Tracking Systems}
+Several researchers are concerned with the validity of studies based on bug tracking systems \cite{Aranda2009,Bird2009,Nguyen2010,Ayari2007}.
+The main concern is the accuracy of the information that is reported in such systems.
+\end{section}
+
+\begin{section}{Background}
+\end{section}
+
+  \begin{subsection}{Bug Tracking Systems}
 Bug workflow. Bugzilla. \cite{Bird2009}
 
-\subsection{The Eclipse Project}
+Bug tracking systems are applications that allow users and developers of a software system to manage a list of bugs for the system.
+Usually, users and developers can report bugs, along with information such as steps to reproduce the bug and the operating system that was used.
+Then, developers choose bugs to fix and can report on the progress of the bug fixing activities, ask for clarification etc.
+
+One important feature of a bug that is recorded on bug tracking systems is its status.
+The status records the progress of the bug fixing activity, and, as such, provides data to software engineering studies.
+Figure \ref{fig:bugzilla} shows each status that can be recorded in Bugzilla, a popular bug tracking system, along with typical transitions between statuses, i.e., the workflow.
+Other systems present comparable status and transitions.
+
+\begin{figure}[ht]
+	\centering
+		\includegraphics[scale=0.4]{bugzilla.png}
+	\caption{Workflow for Bugzilla. Source: \url{http://www.bugzilla.org/docs/2.18/html/lifecycle.html}.}
+	\label{fig:bugzilla}
+\end{figure}
+
+In the simplest cases, a bug is created and receive the status \texttt{UNCONFIRMED} or \texttt{NEW}.
+Next, it is assigned to a developer, and then it is resolved, possibly by fixing it with a patch on the source code.
+The solution is then verified by someone in the quality assurance team.
+If it passes the quality requirements, it is closed.
+If it doesn’t pass the quality requirements or if the solution only partially fixes the problem, the bug is reopened.
+\end{subsection}
+
+  \begin{subsection}{The Eclipse Project}
 History. \\
@@
 
-\subsection{The Netbeans Project}
+The Eclipse Project was created in 2001 by IBM and released into open source in 2004\footnote{\url{http://www.eclipse.org/org/}}, with focus on the creation of a extensible platform that supports the software development lifecycle.
+Today, it is maintained by its open source community, as well as by IBM employees and people from other organizations.
+
+The Eclipse Project consists of many subprojects, including the Platform project and dozens of plugins.
+Although each subproject is maintained by a distinct team and is relatively autonomous, all of them should follow a general, documented process.
+
+All subprojects use Bugzilla for bug reporting, and there are clear rules regarding the process of change in the software\footnote{\url{http://wiki.eclipse.org/Development_Resources/HOWTO/Bugzilla_Use}}.
+Here is an excerpt of a document in the Eclipse web site that explains the bug fixing process\footnote{\url{http://wiki.eclipse.org/Bug_Reporting_FAQ}} (emphasis added):
+
+\begin{quote}
+When a new bug is entered it begins life with a Status of either \textbf{Unconfirmed} for normal users or \textbf{New} for users with commit privileges.
+The bug is typically assigned to the component owner.
+The component owner will usually use a query of Status = Unconfirmed or New and Assigned to = me to browse what is essentially the component's inbox.
+She or he will assign bug reports to developers.
+Email will be generated to the developer (this part is configurable).
+(Note that the act of reassigning a bug to somebody changes its status to New).
+The developer may change unconfirmed bugs to new.
+
+The assigned developer will “accept” the bug which will change its status to \textbf{Assigned}.
+After working on the bug the developer will mark the bug as \textbf{Resolved} and will select a resolution (Fixed, Invalid, Wontfix, Worksforme).
+
+Once a bug is resolved there are still a few states it can transition to.
+How you choose to use these states will be up to individual teams.
+In our example, when a test pass comes up a component owner can build a query that searched for resolved bugs.
+After testing that the fix worked a resolved bug can be transitioned to \textbf{verified}, directly to \textbf{closed}, or in fact, \textbf{reopened}.
+By searching for bugs with a Status of verified and a resolution of fixed developers can come up with release notes.
+A verified/fixed bug can then be transitioned to closed.
+And yes, closed bugs can be reopened if need be.
+As an added bonus other bug status can be transitioned through verified to closed as well.
+This gives developers the opportunity to test worksforme claims.
+\end{quote}
+
+There’s also a recommendation to follow the four eyes principle in the verification process (emphasis added):
+
+\begin{quote}
+When a developer fixes the bug, the status is set to RESOLVED - FIXED, and it is assigned to another developer on the team to verify.
+\textbf{It is important that the verifier be a different person than the fixer} because the fixer is too close to the code and thus may not be as diligent at testing the corner cases.
+\end{quote}
+
+Finally, there are additional guidelines regarding the workflow:
+
+\begin{quote}
+	\begin{itemize}
+\item All bugs should be verified before the next integration build.
+
+\item When a committer verifies a fix, the status is changed to VERIFIED.
+
+\item When the project does a major release, the VERIFIED bugs are changed to CLOSED.
+	\end{itemize}
+\end{quote}
+\end{subsection}
+
+  \begin{subsection}{The Netbeans Project}
 History. \\
 Recommendations for developers.
+\end{subsection}
 
-\section{Experimental Setup}
+\begin{section}{Experimental Setup}
 
@@
 
-\section{Results}
+We have had access to Eclipse Bugzilla database from October, 2001 to June, 2010, in MySQL format.
+The data was made available as part of the Mining Software Repositories 2011 Challenge\footnote{\url{http://www.msrconf.org/msr-challenge.html}}.
 
-\section{Conclusion}
+The database includes data from 155 Eclipse subprojects. However, we have discarded subprojects in which no bug was ever reopened, and subprojects with less than 50 reported bugs.
+
+For the remaining 34 subprojects, we have selected only bugs that were verified until December 2009.
+The rationale is that if, within six months, a verified bug was not reopened, it probably will not be reopened after that and it is, thus, fully fixed.
+\end{section}
+
+\begin{section}{Results}
+
+	O que quer dizer “a solução de um bug foi verificada”?
+
+	Como já foi comentado anteriormente, o Bugzilla permite que os relatórios de bugs sejam marcados como RESOLVED (resolvido) e, depois, como VERIFIED (verificado).
+Mas afinal, o que significa verificar a solução de um bug?
+Que atividades ou checagens foram realizadas antes de se marcar o bug como VERIFIED?
+
+	A documentação do Bugzilla tem algo a dizer sobre isso o status VERIFIED:
+
+	“QA [quality assurance team] has looked at the bug and the resolution and agrees that the appropriate resolution has been taken.”
+
+	Essa definição sugere que a verificação deve ser feita por uma equipe especializada em controle de qualidade.
+Além disso, o verificador deve concordar que a resolução é apropriada.
+A definição do Bugzilla para por aí, mas eu consigo imaginar alguns critérios para considerar uma resolução apropriada:
+
+	O bug não se manifesta após a aplicação da solução.
+	A solução não gera novos bugs.
+	O código da solução segue as convenções de codificação e políticas do projeto.
+	A solução possui uma boa relação custo-benefício — isto é, ou não existem soluções melhores ou a solução é um paliativo por conta da alta complexidade de uma solução melhor.
+	Minha expectativa, portanto, é que, antes de o bug ser marcado como VERIFICADO, ele passe pelas seguintes atividades:
+
+	Aplicação do patch com a solução do bug na versão do software onde o bug foi encontrado.
+	Execução do software com o bug, seguindo os passos para reprodução do bug, e verificando que o bug não se manifesta
+	Inspeção do código-fonte. Deve ser verificada a aderência às convenções de codificação do projeto e a outras políticas (por exemplo, cada patch deve vir com testes de unidade).
+	Execução de testes de unidade (se houver). Deve se verificar que todos os testes passam.
+	Bom, mas isso é a teoria.
+Quais atividades são executadas na prática, em projetos reais, antes de se marcar um bug como VERIFIED?
+Para responder a essa pergunta, analisei um total de 80 bugs, considerando os projetos Eclipse/EMF, Eclipse/Platform, Netbeans/versioncontrol e Netbeans/profiler.
+Eis alguns casos verificados nesses projetos:
+
+	O reporter (ou outro usuário que também enfrenta o problema) verificou que, em uma build do produto que inclui a solução, o bug deixou de se manifestar.
+Aparentemente não há inspeção do código-fonte, e nem uma equipe de QA.
+Situação frequente no Netbeans.
+Pode ser um procedimento necessário quando os desenvolvedores não conseguem reproduzir o bug.
+	A solução está disponível em um build que é divulgado no site do projeto.
+Esse critério é usado pelo projeto Eclipse/EMF e foi decidido em reunião.
+O projeto usa o status FIXED para bugs cuja solução está no CVS.
+	Bugs que já foram resolvidos há muito tempo são marcados como verificados, por se entender que, se nenhuma problema ocorreu até então, isso significa que a solução é adequada.
+	A mesma pessoa que resolve o bug o marca como VERIFIED logo depois. Aparentemente não existe verificação de fato nesse caso.
+	O desenvolvedor que resolve o bug solicita que outro desenvolvedor verifique a solução. Comum no Eclipse/Platform.
+	No bug 249436 do Eclipse/EMF, um patch é submetido por um desenvolvedor, mas outro desenvolvedor acaba reescrevendo o patch, fornecendo evidências de que o código-fonte foi inspecionado.
+	No bug 269789 do Eclipse/EMF, há evidências da elaboração e da execução de testes de unidade.
+	A análise sugere que o status VERIFIED significa coisas diferentes em diferentes projetos.
+A execução de uma versão corrigida do software parece ser o método predominante de verificação.
+Em alguns casos, o status VERIFIED não representa verificação, e sim a indicação de que a solução está disponível para o usuário final.
+Evidências de inspeção de código e execução de testes de unidade só foram encontradas no projeto Eclipse/Platform.
+
+\end{section}
+
+\begin{section}{Conclusion}
+\end{section}
 

### 2011-10-27 contents.tex
Adding more snippets. Fixing wrong latex command.

@@
 Bug workflow. Bugzilla. \cite{Bird2009}
+
+Reposit ́orios de bugs.
+Arquivam relat ́orios de defeitos e solicitac ̧ ̃oes de mudan ̧ca.
+Tais reposito ́rios ajudam os desenvolvedores a registrar, discutir e acompanhar a situa ̧ca ̃o de cada solicita ̧c ̃ao, desde a atribui ̧ca ̃o de uma solicita ̧ca ̃o a um desenvolvedor at ́e a verifica ̧c ̃ao da corre ̧c ̃ao ou melhoria.
+Exemplos incluem o Bugzilla, o Trac e o Jira.
 

### 2012-01-17 contents.tex
Transforming in short paper.

--- 2011-10-27 contents.tex
@@
+% http://wiki.eclipse.org/Bug_Reporting_FAQ
+%   when a test pass comes up a component owner can build a query that searched for resolved bugs
+%   After testing that the fix worked a resolved bug can be transitioned to verified, directly to closed, or in fact, reopened. 
+%   By searching for bugs with a Status of verified and a resolution of fixed developers can come up with release notes.
+%   Severity is assigned by a user and describes the level of impact the bug is having on them. Priority is assigned by the developer and describes the importance a developer places on fixing the bug.
+
 \begin{section}{Introduction}
 
-Software development is a complex activity that demands the coordination of many people, such as programmers, designers, and managers.
-Software development activities should be coordinated in order to deliver products with high quality and on schedule.
-Many practices and processes have been proposed in order to aid the software development, targeting at high quality, low cost, fast delivery, or a combination of these.
+Bug repositories (or bug tracking systems) have since a long time been used in software projects to support coordination among stakeholders.
+Such systems record discussion and progress of software evolution activities, such as corrective, perfective, and ...
+changes \cite{Antoniol2008}.
+Hence, bug repositories are an opportunity to researchers who intend to investigate issues related to the quality of the product and of the process of a software development team, which can give insights about the nature of software development itself.
 
-Several researchers are concerned with the validity of studies based on bug tracking systems \cite{Aranda2009,Bird2009,Nguyen2010,Ayari2007}.
-The main concern is the accuracy of the information that is reported in such systems.
+The information contained in bug repositories have been used in order to predict fault-proneness, to unveil developers' roles from data, to characterize transfer of work in software projects, and so on.
+Also, it has been used to investigate beliefs in software engineering related to software quality, such as the impact of developer turnover and of collective ownership on the occurrence of bugs.
+
+However, mining bug repositories has its own risks.
+Previous research has identified problems of missing data (e.g., rationale, traceability links between reported bug fixes and source code changes) and inaccurate data (e.g., misclassification of bugs) \cite{Aranda2009,Bird2009,Nguyen2010,Ayari2007}.
+
+\TODO why the VERIFIED status is so important?
+
+In summary, for researchers who intend to use data from bug tracking systems, it is not sufficient to understand the data fields in a bug report the way they are documented in the system's documentation.
+The research should also investigate whether the data fields are used often, whether they are used accordingly to the documentation, and even if the data fields are used consistently across the developers.
+
+In this paper, we investigate the status of a bug, a data field present in most bug tracking systems that reports the progress of the bug solving activity.
+In particular, we focus on the activity of verification of a fix.
+We are interested in answering the following exploratory research questions:
+
+\begin{itemize}
+	\item What activity is actually performed when a bug fix is marked as verified?
+	\item When is the verification performed (i.e., just after the fix, or is there a specific period for the verification of fixed bugs?)
+	\item How is the distribution of verification across developers?
+	\item Is there a dedicated team for QA?
+\end{itemize}
+
+For research question xx, qualitative of a small sample. 
+For research question xx, all bugs were considered.
+
 \end{section}
@@
 
-Bug tracking systems are applications that allow users and developers of a software system to manage a list of bugs for the system.
-Usually, users and developers can report bugs, along with information such as steps to reproduce the bug and the operating system that was used.
-Then, developers choose bugs to fix and can report on the progress of the bug fixing activities, ask for clarification etc.
+% Reposit ́orios de bugs.
+Arquivam relat ́orios de defeitos e solicitac ̧ ̃oes de mudan ̧ca.
+Tais reposito ́rios ajudam os desenvolvedores a registrar, discutir e acompanhar a situa ̧ca ̃o de cada solicita ̧c ̃ao, desde a atribui ̧ca ̃o de uma solicita ̧ca ̃o a um desenvolvedor at ́e a verifica ̧c ̃ao da corre ̧c ̃ao ou melhoria.
+Exemplos incluem o Bugzilla, o Trac e o Jira.
+
+Bug tracking systems allow users and developers of a software system to manage a list of bugs for the system.
+Usually, users and developers can report bugs, along with information such as steps to reproduce the bug, its severity, the operating system used etc.
+Developers choose bugs to fix and can report on the progress of the bug fixing activities, ask for clarification etc.
 
@@
 The status records the progress of the bug fixing activity, and, as such, provides data to software engineering studies.
-Figure \ref{fig:bugzilla} shows each status that can be recorded in Bugzilla, a popular bug tracking system, along with typical transitions between statuses, i.e., the workflow.
-Other systems present comparable status and transitions.
+Figure \ref{fig:bugzilla} shows each status that can be recorded in Bugzilla, a popular bug tracking system, along with typical transitions between status values, i.e., the workflow.
+Other systems present comparable status values and transitions.
 
@@
 
-In the simplest cases, a bug is created and receive the status \texttt{UNCONFIRMED} or \texttt{NEW}.
+In the simplest cases, a bug is created and receive the status \texttt{UNCONFIRMED} (if it was created by a regular user) or \texttt{NEW} (if it was created by a developer).
 Next, it is assigned to a developer, and then it is resolved, possibly by fixing it with a patch on the source code.
 The solution is then verified by someone in the quality assurance team.
-If it passes the quality requirements, it is closed.
+If it passes the quality requirements, it is closed and the patch is added to the next release.
 If it doesn’t pass the quality requirements or if the solution only partially fixes the problem, the bug is reopened.
+
 \end{subsection}
+
+\begin{section}{Method}
+
+\end{section}
+
+\begin{section}{Data Extraction}
+
+	We have had access to Eclipse Bugzilla database from October, 2001 to June, 2010, in MySQL format.
+The data was made available as part of the Mining Software Repositories 2011 Challenge\footnote{\url{http://www.msrconf.org/msr-challenge.html}}.
+
+	The database includes data from 155 Eclipse subprojects. However, we have discarded subprojects in which no bug was ever reopened, and subprojects with less than 50 reported bugs.
+
+	For the remaining 34 subprojects, we have selected only bugs that were verified until December 2009.
+The rationale is that if, within six months, a verified bug was not reopened, it probably will not be reopened after that and it is, thus, fully fixed.
 
@@
 
+\end{section}
+
+\begin{section}{Data Sampling}
+	Scripts to classify interesting cases.
+	Random sampling stratified by case.
+
+\end{section}
+
+\begin{section}{Analysis}
+	grounded theory-style.
+
+\end{section}
+
+\begin{section}{Conclusion}
+
+\end{section}
+
 \begin{section}{Experimental Setup}
@@
 
-We have had access to Eclipse Bugzilla database from October, 2001 to June, 2010, in MySQL format.
-The data was made available as part of the Mining Software Repositories 2011 Challenge\footnote{\url{http://www.msrconf.org/msr-challenge.html}}.
-
-The database includes data from 155 Eclipse subprojects. However, we have discarded subprojects in which no bug was ever reopened, and subprojects with less than 50 reported bugs.
-
-For the remaining 34 subprojects, we have selected only bugs that were verified until December 2009.
-The rationale is that if, within six months, a verified bug was not reopened, it probably will not be reopened after that and it is, thus, fully fixed.
 \end{section}

### 2012-01-20 contents.tex
Writing background.

@@
+% http://wiki.eclipse.org/Development_Resources/HOWTO/Bugzilla_Use
+%   The severity tags aren't used much, except to distinguish enhancements from bugs. Typically the users specify severities as they see fit.
+
 % http://wiki.eclipse.org/Bug_Reporting_FAQ
@@
 Bug tracking systems allow users and developers of a software system to manage a list of bugs for the system.
-Usually, users and developers can report bugs, along with information such as steps to reproduce the bug, its severity, the operating system used etc.
-Developers choose bugs to fix and can report on the progress of the bug fixing activities, ask for clarification etc.
+Usually, users and developers can report bugs, along with information such as steps to reproduce the bug, its severity and the operating system used.
+Developers choose bugs to fix and can report on the progress of the bug fixing activities, ask for clarification, discuss causes for the bug etc.
+
+In this research, we focus on Bugzilla, an open source bug tracking system used by notable software projects such as Eclipse, Mozilla, Linux Kernel, Open Office, Apache, and companies such as NASA and Facebook\footnote{Complete list available at http://www.bugzilla.org/installation-list/.}.
+However, the general concepts from Bugzilla should apply to most other bug tracking systems.
 
@@
 The status records the progress of the bug fixing activity, and, as such, provides data to software engineering studies.
-Figure \ref{fig:bugzilla} shows each status that can be recorded in Bugzilla, a popular bug tracking system, along with typical transitions between status values, i.e., the workflow.
-Other systems present comparable status values and transitions.
+Figure \ref{fig:bugzilla} shows each status that can be recorded in Bugzilla, along with typical transitions between status values, i.e., the workflow.
 
@@
 In the simplest cases, a bug is created and receive the status \texttt{UNCONFIRMED} (if it was created by a regular user) or \texttt{NEW} (if it was created by a developer).
-Next, it is assigned to a developer, and then it is resolved, possibly by fixing it with a patch on the source code.
-The solution is then verified by someone in the quality assurance team.
-If it passes the quality requirements, it is closed and the patch is added to the next release.
-If it doesn’t pass the quality requirements or if the solution only partially fixes the problem, the bug is reopened.
+Next, it is ASSIGNED to a developer, and then it is RESOLVED, possibly by fixing it with a patch on the source code.
+The solution is then VERIFIED by someone in the quality assurance team.
+If it passes the quality requirements, it is CLOSED when the next release of the software is released.
+If it does not pass the quality requirements or if the solution only partially fixes the problem, the bug is REOPENED.
+
+In order to understand the bug fixing process of a software team, however, it should be possible to know what activities are performed upon change of status.
+For example, what steps are made before changing the status of a bug to RESOLVED (fixed)?
+Has the patch been applied to the version control system or is it presented as diff file attached to the bug report?
+Has the patched source code been built and made available for end users (e.g., as a nightly build)?
+The documentation for Bugzilla does not prescribe specific activities for marking a bug as RESOLVED, relegating instead this decision to each software project.
+
+Similarly, what do developers do before marking a bug as VERIFIED?
+Bugzilla documentation states that, when a bug is VERIFIED, it means that ``QA [quality assurance team] has looked at the bug and the resolution and agrees that the appropriate resolution has been taken''\cite{\url{https://landfill.bugzilla.org/bugzilla-3.6-branch/page.cgi?id=fields.html}}.
+Again, the definition is loose, although it assumes the existence of a QA role.
+
+The developer documentation for both Eclipse and Netbeans are unclear about how the verification of a bug fix is performed.
+So, what forms of  software verification are they used in the verification process of these projects?
+% In some parts they mention running unit tests, and running a patched version of the software and try to reproduce the bug.
+
+Software verification techniques are classified in static and dynamic [cite Sommervile].
+Static techniques include source code inspection, automated static analysis, and formal verification.
+Dynamic techniques, or testing, involve executing the software system under certain conditions and comparing its actual behavior with the intended behavior.
+Testing can be done in an improvised way (ad hoc testing), or can be compiled in a list of test cases, which can then be automated (e.g., as unit tests).
+
+While mining bug repositories, one cannot assume that the VERIFIED status comprises all forms of software verification.
+Also, s/he cannot rely on the information provided by the bug tracking system or the developer documentation for a project, since it can be unspecific.
+Such information can only be assessed by taking a close look on actual bug reports.
 

### 2012-01-20 contents.tex
Writing Methods section.

@@
 \begin{itemize}
-	\item What activity is actually performed when a bug fix is marked as verified?
-	\item When is the verification performed (i.e., just after the fix, or is there a specific period for the verification of fixed bugs?)
-	\item How is the distribution of verification across developers?
-	\item Is there a dedicated team for QA?
+	\item How is the verification actually performed? Are there performed ad hoc tests, automated tests, code inspection?
+	\item When is the verification performed? Is it performed just after the fix, or is there a specific period for the verification of fixed bugs?
+	\item Who performs the verification? Is there a dedicated team for QA? 
+	% \item How is the distribution of verification across developers?
+	% \item Is there a dedicated team for QA?
 \end{itemize}
@@
 \begin{section}{Background}
-\end{section}
-
-  \begin{subsection}{Bug Tracking Systems}
-Bug workflow. Bugzilla. \cite{Bird2009}
+% Bug workflow. Bugzilla. \cite{Bird2009}
 
@@
 
+\TODO Roles: reporter, fixer, verifier, reopener, closer.
+
 In order to understand the bug fixing process of a software team, however, it should be possible to know what activities are performed upon change of status.
@@
 Similarly, what do developers do before marking a bug as VERIFIED?
-Bugzilla documentation states that, when a bug is VERIFIED, it means that ``QA [quality assurance team] has looked at the bug and the resolution and agrees that the appropriate resolution has been taken''\cite{\url{https://landfill.bugzilla.org/bugzilla-3.6-branch/page.cgi?id=fields.html}}.
+Bugzilla documentation states that, when a bug is VERIFIED, it means that ``QA [quality assurance team] has looked at the bug and the resolution and agrees that the appropriate resolution has been taken''\footnote{\url{https://landfill.bugzilla.org/bugzilla-3.6-branch/page.cgi?id=fields.html}}.
 Again, the definition is loose, although it assumes the existence of a QA role.
 
-The developer documentation for both Eclipse and Netbeans are unclear about how the verification of a bug fix is performed.
+The developer documentation for both Eclipse and NetBeans are unclear about how the verification of a bug fix is performed.
 So, what forms of  software verification are they used in the verification process of these projects?
@@
 
-\end{subsection}
+% O bug não se manifesta após a aplicação da solução.
+% A solução não gera novos bugs.
+% O código da solução segue as convenções de codificação e políticas do projeto.
+% A solução possui uma boa relação custo-benefício — isto é, ou não existem soluções melhores ou a solução é um paliativo por conta da alta complexidade de uma solução melhor.
+% 
+% Minha expectativa, portanto, é que, antes de o bug ser marcado como VERIFICADO, ele passe pelas seguintes atividades:
+% 
+% Aplicação do patch com a solução do bug na versão do software onde o bug foi encontrado.
+% Execução do software com o bug, seguindo os passos para reprodução do bug, e verificando que o bug não se manifesta
+% Inspeção do código-fonte. Deve ser verificada a aderência às convenções de codificação do projeto e a outras políticas (por exemplo, cada patch deve vir com testes de unidade).
+% Execução de testes de unidade (se houver). Deve se verificar que todos os testes passam.
+
+\end{section}
 
@@
 
-\end{section}
+	% \item How is the verification actually performed?
+	% \item When is the verification performed (i.e., just after the fix, or is there a specific period for the verification of fixed bugs?)
+	% \item Who performs the verification?
 
-\begin{section}{Data Extraction}
+In order to answer our exploratory research questions, we followed the following method:
 
-	We have had access to Eclipse Bugzilla database from October, 2001 to June, 2010, in MySQL format.
-The data was made available as part of the Mining Software Repositories 2011 Challenge\footnote{\url{http://www.msrconf.org/msr-challenge.html}}.
+\begin{enumerate}
+	\item \textbf{Data extraction}: we have obtained publicly available raw data from the Bugzilla repositories of two popular integrated development environments, Eclipse and Bugzilla.
 
-	The database includes data from 155 Eclipse subprojects. However, we have discarded subprojects in which no bug was ever reopened, and subprojects with less than 50 reported bugs.
+	\item \textbf{Data sampling}: we have chosen two subprojects. After that, we sampled bugs with interesting characteristics.
 
-	For the remaining 34 subprojects, we have selected only bugs that were verified until December 2009.
-The rationale is that if, within six months, a verified bug was not reopened, it probably will not be reopened after that and it is, thus, fully fixed.
+	\item \textbf{Data analysis}: we have read the full bug reported for the sampled bugs. Also, we have performed quantitative analysis on all bugs.
 
-\begin{subsection}{The Eclipse Project}
-	History. \\
-	Recommendations about Bugzilla usage. Four-eyes principle.
+	\item ...
+\end{enumerate}
 
-	The Eclipse Project was created in 2001 by IBM and released into open source in 2004\footnote{\url{http://www.eclipse.org/org/}}, with focus on the creation of a extensible platform that supports the software development lifecycle.
-Today, it is maintained by its open source community, as well as by IBM employees and people from other organizations.
+\begin{subsection}{Available Data}
 
-	The Eclipse Project consists of many subprojects, including the Platform project and dozens of plugins.
-Although each subproject is maintained by a distinct team and is relatively autonomous, all of them should follow a general, documented process.
+	In order to perform the desired analysis, we needed access to all data recorded by Bugzilla for a specific project.
+We have found such data for two projects --- Eclipse and NetBeans --- from the domain of integrated development environments.
+The data were made available as part of the Mining Software Repositories 2011 Challenge\footnote{\url{http://www.msrconf.org/msr-challenge.html}} in the form of MySQL database dumps.
+The files contain all tables from the respective databases, except for the tables containing information about developers.
 
-	All subprojects use Bugzilla for bug reporting, and there are clear rules regarding the process of change in the software\footnote{\url{http://wiki.eclipse.org/Development_Resources/HOWTO/Bugzilla_Use}}.
-Here is an excerpt of a document in the Eclipse web site that explains the bug fixing process\footnote{\url{http://wiki.eclipse.org/Bug_Reporting_FAQ}} (emphasis added):
+	% We have had access to Eclipse Bugzilla database from October, 2001 to June, 2010, and to NetBeans Bugzilla database from ...
+to ....
+The data was made available as part of the Mining Software Repositories 2011 Challenge\footnote{\url{http://www.msrconf.org/msr-challenge.html}} as a SQL dump.
+The SQL dump contains all information recorded by Bugzilla for the projects (except for information about developers, which was removed for privacy reasons).
 
-	\begin{quote}
-	When a new bug is entered it begins life with a Status of either \textbf{Unconfirmed} for normal users or \textbf{New} for users with commit privileges.
-The bug is typically assigned to the component owner.
-The component owner will usually use a query of Status = Unconfirmed or New and Assigned to = me to browse what is essentially the component's inbox.
-She or he will assign bug reports to developers.
-Email will be generated to the developer (this part is configurable).
-(Note that the act of reassigning a bug to somebody changes its status to New).
-The developer may change unconfirmed bugs to new.
+	Eclipse development began in late 1998 with IBM\footnote{\url{http://www.ibm.com/developerworks/rational/library/nov05/cernosek/}}.
+It was licensed as open source in November, 2001.
+In 2004, IBM created the Eclipse Foundation in order to reduce the perception of Eclipse as a IBM-controlled project.
 
-	The assigned developer will “accept” the bug which will change its status to \textbf{Assigned}.
-After working on the bug the developer will mark the bug as \textbf{Resolved} and will select a resolution (Fixed, Invalid, Wontfix, Worksforme).
+	NetBeans\footnote{\url{http://netbeans.org/about/history.html}} started  as a student project in 1996, was bought by Sun Microsystems in Oct, 1999, and was open sourced in Mar, 2000.
 
-	Once a bug is resolved there are still a few states it can transition to.
-How you choose to use these states will be up to individual teams.
-In our example, when a test pass comes up a component owner can build a query that searched for resolved bugs.
-After testing that the fix worked a resolved bug can be transitioned to \textbf{verified}, directly to \textbf{closed}, or in fact, \textbf{reopened}.
-By searching for bugs with a Status of verified and a resolution of fixed developers can come up with release notes.
-A verified/fixed bug can then be transitioned to closed.
-And yes, closed bugs can be reopened if need be.
-As an added bonus other bug status can be transitioned through verified to closed as well.
-This gives developers the opportunity to test worksforme claims.
-	\end{quote}
+	Both Eclipse and NetBeans are composed of many subprojects.
+The data comprises all bug reports from all Eclipse subprojects from October, 2001 to June, 2010, and from all NetBeans subprojects from June, 1998 to June, 2010.
 
-	There’s also a recommendation to follow the four eyes principle in the verification process (emphasis added):
+	There 316,911 bug reports for 155 subprojects in Eclipse, and 185,578 bug reports for 39 subprojects in NetBeans.
 
-	\begin{quote}
-	When a developer fixes the bug, the status is set to RESOLVED - FIXED, and it is assigned to another developer on the team to verify.
-\textbf{It is important that the verifier be a different person than the fixer} because the fixer is too close to the code and thus may not be as diligent at testing the corner cases.
-	\end{quote}
-
-	Finally, there are additional guidelines regarding the workflow:
-
-	\begin{quote}
-		\begin{itemize}
-	\item All bugs should be verified before the next integration build.
-
-	\item When a committer verifies a fix, the status is changed to VERIFIED.
-
-	\item When the project does a major release, the VERIFIED bugs are changed to CLOSED.
-		\end{itemize}
-	\end{quote}
 \end{subsection}
 
-\begin{subsection}{The Netbeans Project}
-	History. \\
-	Recommendations for developers.
+\begin{subsection}{Sampling of Bug Reports}
+
+	In order to obtain insight about the research questions, we read a sample of bug reports from both projects.
+For each project, two subprojects were chosen randomly among the projects with most verified bugs, resulting in the following subprojects: Eclipse/Platform, Eclipse/EMF, Netbeans/profiler, Netbeans/versioncontrol.
+We have sampled 20 bug reports for each subproject, totalizing 80 bug reports.
+
+\end{subsection}
+
+\begin{subsection}{Analysis}
+
+	Each of the sampled bugs was read.
+
 \end{subsection}
@@
 
-\begin{section}{Data Sampling}
-	Scripts to classify interesting cases.
-	Random sampling stratified by case.
+\begin{section}{Results and Discussion}
+	Some insights were gained in this step:
 
-\end{section}
-
-\begin{section}{Analysis}
-	grounded theory-style.
+	\begin{itemize}
+		\item In some projects, the status VERIFIED does not correspond to software verification.
+In Eclipse/EMF, for example, it means that the fix is available in a build that is promoted on the project website\footnote{\url{http://wiki.eclipse.org/Modeling_PMC_Meeting,_2007-10-16}}.
+		\item In the NetBeans subprojects that were analyzed, it was common that the reporter also verified the fix. In some bugs, this was necessary since the developer was not able to reproduce the bug.
+		\item Often, the fixer verifies the bug just after marking it as fixed.
+		\item Sometimes, several old bugs are marked as VERIFIED at the same time.
+		\item In Eclipse/Platform, it is common that the fixer requests someone else to verify the fix.
+		\item In Eclipse/EMF, bug #249436, one developer rewrites the patch submitted by other developer.
+		\item In Eclipse/EMF, bug #269789, a test case is used to assess the proposed fix.
+	\end{itemize}
 
@@
 
-\begin{section}{Experimental Setup}
-
-\begin{verbatim}
-Data extraction
-Data processing: 
-- filtering (bugs up to december 20xx)
-- classification script
-Sampling:
-- random sampling (for coding)
-- entire data set
-Analysis:
-- grounded theory (open coding)
-- latent Dirichlet allocation
-- descriptive statistics
-\end{verbatim}
-
-\end{section}
-
-\begin{section}{Results}
-
-	O que quer dizer “a solução de um bug foi verificada”?
-
-	Como já foi comentado anteriormente, o Bugzilla permite que os relatórios de bugs sejam marcados como RESOLVED (resolvido) e, depois, como VERIFIED (verificado).
-Mas afinal, o que significa verificar a solução de um bug?
-Que atividades ou checagens foram realizadas antes de se marcar o bug como VERIFIED?
-
-	A documentação do Bugzilla tem algo a dizer sobre isso o status VERIFIED:
-
-	“QA [quality assurance team] has looked at the bug and the resolution and agrees that the appropriate resolution has been taken.”
-
-	Essa definição sugere que a verificação deve ser feita por uma equipe especializada em controle de qualidade.
-Além disso, o verificador deve concordar que a resolução é apropriada.
-A definição do Bugzilla para por aí, mas eu consigo imaginar alguns critérios para considerar uma resolução apropriada:
-
-	O bug não se manifesta após a aplicação da solução.
-	A solução não gera novos bugs.
-	O código da solução segue as convenções de codificação e políticas do projeto.
-	A solução possui uma boa relação custo-benefício — isto é, ou não existem soluções melhores ou a solução é um paliativo por conta da alta complexidade de uma solução melhor.
-	Minha expectativa, portanto, é que, antes de o bug ser marcado como VERIFICADO, ele passe pelas seguintes atividades:
-
-	Aplicação do patch com a solução do bug na versão do software onde o bug foi encontrado.
-	Execução do software com o bug, seguindo os passos para reprodução do bug, e verificando que o bug não se manifesta
-	Inspeção do código-fonte. Deve ser verificada a aderência às convenções de codificação do projeto e a outras políticas (por exemplo, cada patch deve vir com testes de unidade).
-	Execução de testes de unidade (se houver). Deve se verificar que todos os testes passam.
-	Bom, mas isso é a teoria.
-Quais atividades são executadas na prática, em projetos reais, antes de se marcar um bug como VERIFIED?
-Para responder a essa pergunta, analisei um total de 80 bugs, considerando os projetos Eclipse/EMF, Eclipse/Platform, Netbeans/versioncontrol e Netbeans/profiler.
-Eis alguns casos verificados nesses projetos:
-
-	O reporter (ou outro usuário que também enfrenta o problema) verificou que, em uma build do produto que inclui a solução, o bug deixou de se manifestar.
-Aparentemente não há inspeção do código-fonte, e nem uma equipe de QA.
-Situação frequente no Netbeans.
-Pode ser um procedimento necessário quando os desenvolvedores não conseguem reproduzir o bug.
-	A solução está disponível em um build que é divulgado no site do projeto.
-Esse critério é usado pelo projeto Eclipse/EMF e foi decidido em reunião.
-O projeto usa o status FIXED para bugs cuja solução está no CVS.
-	Bugs que já foram resolvidos há muito tempo são marcados como verificados, por se entender que, se nenhuma problema ocorreu até então, isso significa que a solução é adequada.
-	A mesma pessoa que resolve o bug o marca como VERIFIED logo depois. Aparentemente não existe verificação de fato nesse caso.
-	O desenvolvedor que resolve o bug solicita que outro desenvolvedor verifique a solução. Comum no Eclipse/Platform.
-	No bug 249436 do Eclipse/EMF, um patch é submetido por um desenvolvedor, mas outro desenvolvedor acaba reescrevendo o patch, fornecendo evidências de que o código-fonte foi inspecionado.
-	No bug 269789 do Eclipse/EMF, há evidências da elaboração e da execução de testes de unidade.
-	A análise sugere que o status VERIFIED significa coisas diferentes em diferentes projetos.
-A execução de uma versão corrigida do software parece ser o método predominante de verificação.
-Em alguns casos, o status VERIFIED não representa verificação, e sim a indicação de que a solução está disponível para o usuário final.
-Evidências de inspeção de código e execução de testes de unidade só foram encontradas no projeto Eclipse/Platform.
-
-\end{section}
-
-\begin{section}{Conclusion}
-\end{section}
-
 \section*{Acknowledgment}

### 2012-01-25 contents.tex
Adding data.

@@
 		\item Sometimes, several old bugs are marked as VERIFIED at the same time.
-		\item In Eclipse/Platform, it is common that the fixer requests someone else to verify the fix.
+		\item In Eclipse/Platform, it is common that the fixer requests someone else to verify the fix (``Please verify, Tom'').
 		\item In Eclipse/EMF, bug #249436, one developer rewrites the patch submitted by other developer.
@@
 	\end{itemize}
+
+\begin{subsection}{RQ1: How Are Bugs Verified?}
+
+	We analyzed cases in which the bug is fixed, the fixer asks someone else to verify the fix, and then the bug is reopened.
+
+	Findings: ...
+
+\end{subsection}
+
+\begin{subsection}{RQ2: When Are Bugs Verified?}
+
+\end{subsection}
+
+\begin{subsection}{RQ3: Who Verifies Bugs?}
+
+	We counted the number of times each developer verified or fixed a bug.
+We consider that, if a developer is primarily concerned with quality assurance, then s/he verifies much more bugs that s/he fixes.
+
+	We defined ratio as ...
+
+	The verified/fixed ratio for the four projects is shown on Table XXX.
+
+\begin{verbatim}
+	*** eclipse-platform 
+	Verifiers: 4(2.38095238095238%) = 1.08523937083591% of verifications
+	Fixers: 30 -- 17.8571428571429%
+	Total developers: 168
+
+	*** eclipse-emf 
+	Verifiers: 0(0%) = 0% of verifications
+	Fixers: 4 -- 30.7692307692308%
+	Total developers: 13
+
+	*** netbeans-profiler 
+	Verifiers: 3(21.4285714285714%) = 89.9688958009331% of verifications
+	Fixers: 2 -- 14.2857142857143%
+	Total developers: 14
+
+	*** netbeans-versioncontrol 
+	Verifiers: 5(20.8333333333333%) = 93.2291666666667% of verifications
+	Fixers: 6 -- 25%
+	Total developers: 24
+\end{verbatim}
+
+	Clearly, in NetBeans the ``verifier'' role is more pronounced.
+In the subprojects analyzed, approximately 20\% of the developers focus on the verification activity, and they perform about 90\% of the verifications.
+
+	In the Eclipse subprojects, however, such role is not so clear.
+In EMF, there are no ``verifiers''.
+That does not mean much because, as we discovered earlier, the VERIFIED status does not correspond to the verification activity.
+In the case of Eclipse/Platform, only 2\% of the developers are verifiers, and they contribute only to 1\% of the verifications.
+
+\end{subsection}
 

### 2012-01-26 contents.tex
Revision after meeting with Christina. Citing 4-eyes report as motivation and future work.

@@
 
-Bug repositories (or bug tracking systems) have since a long time been used in software projects to support coordination among stakeholders.
+Bug repositories (or bug tracking systems) have for a long time been used in software projects to support coordination among stakeholders.
 Such systems record discussion and progress of software evolution activities, such as corrective, perfective, and ...
 changes \cite{Antoniol2008}.
-Hence, bug repositories are an opportunity to researchers who intend to investigate issues related to the quality of the product and of the process of a software development team, which can give insights about the nature of software development itself.
+Hence, bug repositories are an opportunity to researchers who intend to investigate issues related to the quality of the product and of the process of software development team.
+%, or even issues regarding the nature of software development itself.
 
-The information contained in bug repositories have been used in order to predict fault-proneness, to unveil developers' roles from data, to characterize transfer of work in software projects, and so on.
+The information contained in bug repositories has been used in order to predict fault-proneness, to unveil developers' roles from data, to characterize transfer of work in software projects, and so on.
 Also, it has been used to investigate beliefs in software engineering related to software quality, such as the impact of developer turnover and of collective ownership on the occurrence of bugs.
@@
 
-\TODO why the VERIFIED status is so important?
+  In previous research [cite report], we tried to assess the impact of independent verification of bug fixes on software quality, by mining data from bug repositories. 
+  We relied on reported verifications tasks, as recorded in bug reports, and interpreted the recorded data according to the documentation for the specific bug tracking system used.
+  As the partial results suggested that verification has no impact on software quality, we questioned the accuracy of the data about verification of bug fixes, ad thus decided to investigate how verification is actually performed and reported on specific software projects.
 
-In summary, for researchers who intend to use data from bug tracking systems, it is not sufficient to understand the data fields in a bug report the way they are documented in the system's documentation.
-The research should also investigate whether the data fields are used often, whether they are used accordingly to the documentation, and even if the data fields are used consistently across the developers.
-
-In this paper, we investigate the status of a bug, a data field present in most bug tracking systems that reports the progress of the bug solving activity.
-In particular, we focus on the activity of verification of a fix.
-We are interested in answering the following exploratory research questions:
+	In particular, we are interested in answer the following exploratory research questions for a limited set of projects:
 
 	\begin{itemize}
-	\item How is the verification actually performed? Are there performed ad hoc tests, automated tests, code inspection?
-	\item When is the verification performed? Is it performed just after the fix, or is there a specific period for the verification of fixed bugs?
-	\item Who performs the verification? Is there a dedicated team for QA? 
+		\item \textbf{How} is the verification performed: are there performed ad hoc tests, automated tests, code inspection?
+		\item \textbf{Who} performs the verification: is there a dedicated team for QA? 
+		\item \textbf{When} is the verification performed: is it performed just after the fix, or is there a verification phase?
 		% \item How is the distribution of verification across developers?
@@
 
-For research question xx, qualitative of a small sample. 
-For research question xx, all bugs were considered.
+% \TODO why the VERIFIED status is so important?
+
+% In summary, for researchers who intend to use data from bug tracking systems, it is not sufficient to understand the data fields in a bug report the way they are documented in the system's documentation.
+The research should also investigate whether the data fields are used often, whether they are used accordingly to the documentation, and even if the data fields are used consistently across the developers.
+
+% In this paper, we investigate the status of a bug, a data field present in most bug tracking systems that reports the progress of the bug solving activity.
+In particular, we focus on the activity of verification of a fix.
+We are interested in answering the following exploratory research questions:
+
+% For research question xx, qualitative of a small sample. 
+% For research question xx, all bugs were considered.
 
@@
 
-Bug tracking systems allow users and developers of a software system to manage a list of bugs for the system.
+Bug tracking systems allow users and developers of a software project to manage a list of bugs for the project.
 Usually, users and developers can report bugs, along with information such as steps to reproduce the bug, its severity and the operating system used.
@@
 In this research, we focus on Bugzilla, an open source bug tracking system used by notable software projects such as Eclipse, Mozilla, Linux Kernel, Open Office, Apache, and companies such as NASA and Facebook\footnote{Complete list available at http://www.bugzilla.org/installation-list/.}.
-However, the general concepts from Bugzilla should apply to most other bug tracking systems.
+The general concepts from Bugzilla should apply to most other bug tracking systems.
 
@@
 If it passes the quality requirements, it is CLOSED when the next release of the software is released.
-If it does not pass the quality requirements or if the solution only partially fixes the problem, the bug is REOPENED.
+If it does not pass the quality requirements or if the solution only partially fixes the problem, the bug is REOPENed.
 
@@
 In order to understand the bug fixing process of a software team, however, it should be possible to know what activities are performed upon change of status.
-For example, what steps are made before changing the status of a bug to RESOLVED (fixed)?
-Has the patch been applied to the version control system or is it presented as diff file attached to the bug report?
-Has the patched source code been built and made available for end users (e.g., as a nightly build)?
+For example, what steps are made before changing the status of a bug to RESOLVED (FIXED)?
+Is the patch applied to the version control system?
+Is it presented as diff file attached to the bug report?
+Is there a build available to end users containing the patch (e.g., a nightly build)?
 The documentation for Bugzilla does not prescribe specific activities for marking a bug as RESOLVED, relegating instead this decision to each software project.
@@
 Bugzilla documentation states that, when a bug is VERIFIED, it means that ``QA [quality assurance team] has looked at the bug and the resolution and agrees that the appropriate resolution has been taken''\footnote{\url{https://landfill.bugzilla.org/bugzilla-3.6-branch/page.cgi?id=fields.html}}.
-Again, the definition is loose, although it assumes the existence of a QA role.
+Again, the definition is broad, although it assumes the existence of a QA team.
 
@@
 While mining bug repositories, one cannot assume that the VERIFIED status comprises all forms of software verification.
-Also, s/he cannot rely on the information provided by the bug tracking system or the developer documentation for a project, since it can be unspecific.
+Also, one cannot rely on the information provided by the bug tracking system or the developer documentation for a project, since it can be too generic to be useful.
 Such information can only be assessed by taking a close look on actual bug reports.
@@
 	In order to perform the desired analysis, we needed access to all data recorded by Bugzilla for a specific project.
-We have found such data for two projects --- Eclipse and NetBeans --- from the domain of integrated development environments.
+We have found such data for two projects---Eclipse and NetBeans---from the domain of integrated development environments.
 The data were made available as part of the Mining Software Repositories 2011 Challenge\footnote{\url{http://www.msrconf.org/msr-challenge.html}} in the form of MySQL database dumps.
@@
 
-	NetBeans\footnote{\url{http://netbeans.org/about/history.html}} started  as a student project in 1996, was bought by Sun Microsystems in Oct, 1999, and was open sourced in Mar, 2000.
+	NetBeans\footnote{\url{http://netbeans.org/about/history.html}} started  as a student project in 1996. It was bought by Sun Microsystems in October, 1999, and it was open sourced in March, 2000.
 
@@
 
-	There 316,911 bug reports for 155 subprojects in Eclipse, and 185,578 bug reports for 39 subprojects in NetBeans.
+	There are 316,911 bug reports for 155 subprojects in Eclipse, and 185,578 bug reports for 39 subprojects in NetBeans.
 
@@
 
-\end{subsection}
-
-\begin{subsection}{Analysis}
-
-	Each of the sampled bugs was read.
+	\TODO this sampling was made just for the preliminary, qualitative analysis. Maybe we should describe preliminary analysis as an unstructured inquiry (instead of detailing the methods for it).
 
@@
 \begin{section}{Results and Discussion}
-	Some insights were gained in this step:
+
+\begin{subsection}{Preliminary Analysis}
+	Some insights were gained in this step, partially answering the who, when, and how questions:
 
 	\begin{itemize}
-		\item In some projects, the status VERIFIED does not correspond to software verification.
+		\item [how] In some projects, the status VERIFIED does not correspond to software verification.
 In Eclipse/EMF, for example, it means that the fix is available in a build that is promoted on the project website\footnote{\url{http://wiki.eclipse.org/Modeling_PMC_Meeting,_2007-10-16}}.
-		\item In the NetBeans subprojects that were analyzed, it was common that the reporter also verified the fix. In some bugs, this was necessary since the developer was not able to reproduce the bug.
-		\item Often, the fixer verifies the bug just after marking it as fixed.
-		\item Sometimes, several old bugs are marked as VERIFIED at the same time.
-		\item In Eclipse/Platform, it is common that the fixer requests someone else to verify the fix (``Please verify, Tom'').
-		\item In Eclipse/EMF, bug #249436, one developer rewrites the patch submitted by other developer.
-		\item In Eclipse/EMF, bug #269789, a test case is used to assess the proposed fix.
+		\item [(who)] In the NetBeans subprojects that were analyzed, it was common that the reporter also verified the fix.
+In some bugs, this was necessary since the developer was not able to reproduce the bug.
+		\item [(who, when)] Often, the fixer verifies the bug just after marking it as fixed.
+		\item [(when)] Sometimes, several old bugs are marked as VERIFIED at the same time.
+		\item [(how?)] In Eclipse/Platform, it is common that the fixer requests someone else to verify the fix (``Please verify, Tom'').
+		\item [(how)] In Eclipse/EMF, bug #249436, one developer rewrites the patch submitted by other developer.
+		\item [(how)] In Eclipse/EMF, bug #269789, a test case is used to assess the proposed fix.
 	\end{itemize}
+\end{subsection}
 
@@
 
-	Clearly, in NetBeans the ``verifier'' role is more pronounced.
-In the subprojects analyzed, approximately 20\% of the developers focus on the verification activity, and they perform about 90\% of the verifications.
+	In NetBeans it is possible to infer the existence of a QA team, composed by approximately 20\% of the developers, that performs about 90\% of all verifications.
 
-	In the Eclipse subprojects, however, such role is not so clear.
-In EMF, there are no ``verifiers''.
-That does not mean much because, as we discovered earlier, the VERIFIED status does not correspond to the verification activity.
-In the case of Eclipse/Platform, only 2\% of the developers are verifiers, and they contribute only to 1\% of the verifications.
+	In the Eclipse subprojects, there is no evidence of a dedicated QA team.
+In Eclipse/EMF no developer focus specifically on verification (it should be remembered that verification in EMF has a peculiar meaning).
+In Eclipse/Platform, only 2\% of the developers focus on verification tasks, however, they contribute only to 1\% of the verifications.
 
@@
 \begin{section}{Conclusion}
+  The VERIFIED status is used in a variety of distinct ways, not always consistent with the traditional definition of software verification ...
+
+	With the knowledge obtained from this exploratory research, we can improve and extend our previous work on the impact of independent verification on software quality. Examples of refined questions:
+
+	\begin{itemize}
+		\item Does independent verification improve software quality when it is performed by developers from the QA team?
+		\item Does independent verification improve software quality when it is performed as an separate phase of the software release cycle?
+		\item Does independent verification improve software quality when a particular verification technique (e.g., inspections, unit testing) is used?
+	\end{itemize}
 

### 2012-01-30 contents.tex
Answering the question: when are bug fixes verified?

@@
 
+  \TODO the method has changed. Please ignore this section.
+
 In order to answer our exploratory research questions, we followed the following method:
@@
 	In order to obtain insight about the research questions, we read a sample of bug reports from both projects.
-For each project, two subprojects were chosen randomly among the projects with most verified bugs, resulting in the following subprojects: Eclipse/Platform, Eclipse/EMF, Netbeans/profiler, Netbeans/versioncontrol.
+For each project, two subprojects were chosen randomly among the projects with most verified bugs, resulting in the following subprojects: Eclipse/Platform, Eclipse/EMF, Netbeans/platform, Netbeans/versioncontrol.
 We have sampled 20 bug reports for each subproject, totalizing 80 bug reports.
@@
 		\item [how] In some projects, the status VERIFIED does not correspond to software verification.
-In Eclipse/EMF, for example, it means that the fix is available in a build that is promoted on the project website\footnote{\url{http://wiki.eclipse.org/Modeling_PMC_Meeting,_2007-10-16}}.
+In Eclipse/EMF, for example, it means that the fix is available in a build that is promoted on the project website\footnote{\url{http://wiki.eclipse.org/Modeling_PMC_Meeting,_2007-10-16} (between releases 3.3.1 and 3.3.1.1)}.
+Further evidence: bug #254489 ``Please do not change the state of the bugs since the status verified fixed should be only there when we publish a new build.
+:-)''
 		\item [(who)] In the NetBeans subprojects that were analyzed, it was common that the reporter also verified the fix.
@@
 		\item [(when)] Sometimes, several old bugs are marked as VERIFIED at the same time.
-		\item [(how?)] In Eclipse/Platform, it is common that the fixer requests someone else to verify the fix (``Please verify, Tom'').
+		\item [(how?)] In both Eclipse and NetBeans, it is common that the fixer requests someone else to verify the fix (``Please verify, Tom'').
 		\item [(how)] In Eclipse/EMF, bug #249436, one developer rewrites the patch submitted by other developer.
 		\item [(how)] In Eclipse/EMF, bug #269789, a test case is used to assess the proposed fix.
+		\item [(how)] Eclipse/Platform, bug #210533: ``Verified by code inspection''.
 	\end{itemize}
+
+	\TODO: ``Verified in/with [build]'', ``Verified by/through [verification technique]''.
+
 \end{subsection}
@@
 \begin{subsection}{RQ2: When Are Bugs Verified?}
+
+	In order to answer whether there is a verification phase for the projects analyzed, we plotted the cumulative number of verifications along the lifetime of each project, as shown in Figure XXX, in red.
+The vertical lines represent release dates for each version.
+As a basis for comparison, the number of bug fixes is also plotted, in black.
+
+	We found evidence of a verification phase for Eclipse/Platform.
+In this project, verifications appear to be performed in bursts, with periods of intense activity separated by about 6 to 7 weeks of low or no activity.
+By taking a closer look in the period between the releases 3.5 and 3.6, it is clear that the periods of intense verification activity occur just before the milestones leading to 3.6.
+
+	No such pattern was found for Netbeans' subprojects.
+
+	% We did not find evidence of a verification phase preceding each release.
+Unlike the bug fixing activity, the pace of verification is not steady, either.
+Instead, verification appears to be performed in bursts, which a couple of weeks with low or no activity, followed by a couple of days of intense verification activity.
+
+	% In Eclipse, milestones are roughly 6 weeks apart from each other.
+Milestones are considered good enough for end users\footenote{\url{http://eclipsecon.org/2005/presentations/econ2005-eclipse-way.pdf}} The milestones are, in general, marked by a steep accent in verification just before the milestone date.
+	% Weekly integration builds. Nightly builds. Build quality verified by automatic unit tests. 20k unit tests.
+
+	%%%%%%%%%%%%%%%%%%%%%%
+
+	The first thing that comes to attention are ``cliff walls'', points in time when a large amount of bugs are verified, represented by steep, almost vertical portions of the plotted line.
+They appear in Eclipse/EMF and Netbeans/Platform.
+In Netbeans/Platform, all verifications in the cliff wall preceding version 5.5 share the same message: ``Verification of old issues''.
+For Eclipse/EMF, the message is ``Move to verified as per bug 206558'', which refers to the task of changing bugs marked as FIXED to VERIFIED.
+
+	Cliff walls represent a change in the development process, and the mass verification is needed in order to leverage useful features from Bugzilla (e.g., looking for verified bugs in order to aid the writing of release notes).
+
+	Because of cliff walls, extra care must be taken before doing any analysis that relies on the VERIFIED status.
+In such situations, the VERIFIED status is applied blindly, with no software verification being actually performed.
+Because cliff walls contain a large amount of bugs, they are likely to bias the results of analyses.
+
+	%%%%%%%%%%%%%%%%%%%%%%
+
+	% Some projects have ``cliff walls'' [reference graphic].
+Taking a closer look, the VERIFIED status was used sparingly in Eclipse/EMF, but, suddenly, a huge number of bugs is verified and, after that, bugs tend to be verified frequently.
+It represents a change in the development process, and the mass verification is needed in order to leverage useful features from Bugzilla (e.g., looking for verified bugs in order to aid the writing of release notes).
+
+	% If one is to build a predictive model based on verifications, the cliff walls should be ignored, as they represent unusual activity that has no connection with software verification whatsoever.
+
+	% By visualizing release dates, we tried to assess if there is such a thing as a ``verification phase'', i.e., if verifications are frequent only on a period before the release.
+There was no evidence of such phase on Eclipse subprojects (TODO: look for patterns in NetBeans).
+
+	% However, we found also that, different to what occur with fixes, verifications are not steady, constant.
+In Eclipse/Platform, for instance, there are periods of 3 to 5 weeks of few verifications, followed by a couple of days with intense verification activity (``mini cliff walls'', if you wish...)
+	% 
+	% The verifications of NetBeans/versioncontrol are more steady, continuous. Also, they seem to reflect well the FIXED curve.
+
+	% TODO (future): are verifications more common on mondays on netbeans-versioncontrol?
+
+	% TODO (future): how much time does it take between RESOLVED (FIXED) and VERIFIED?
 
@@
 *** eclipse-platform 
-Verifiers: 4(2.38095238095238%) = 1.08523937083591% of verifications
-Fixers: 30 -- 17.8571428571429%
+Verifiers: 4 (2.4%)
+  1.1% of verifications
+Fixers: 30 (17.9%)
 Total developers: 168
@@
 *** eclipse-emf 
-Verifiers: 0(0%) = 0% of verifications
-Fixers: 4 -- 30.7692307692308%
+Verifiers: 0 (0.0%)
+  0.0% of verifications
+Fixers: 4 (30.8%)
 Total developers: 13
 
-*** netbeans-profiler 
-Verifiers: 3(21.4285714285714%) = 89.9688958009331% of verifications
-Fixers: 2 -- 14.2857142857143%
-Total developers: 14
+*** netbeans-platform 
+Verifiers: 5 (21.7%)
+  93.3% of verifications
+Fixers: 6 (26.1%)
+Total developers: 23
 
 *** netbeans-versioncontrol 
-Verifiers: 5(20.8333333333333%) = 93.2291666666667% of verifications
-Fixers: 6 -- 25%
+Verifiers: 5 (20.8%)
+  93.2% of verifications
+Fixers: 6 (25.0%)
 Total developers: 24 
@@
 	In the Eclipse subprojects, there is no evidence of a dedicated QA team.
-In Eclipse/EMF no developer focus specifically on verification (it should be remembered that verification in EMF has a peculiar meaning).
+In Eclipse/EMF no developer focus specifically on verification.
 In Eclipse/Platform, only 2\% of the developers focus on verification tasks, however, they contribute only to 1\% of the verifications.
+
+% (it should be remembered that verification in EMF has a peculiar meaning)
 

### 2012-02-02 contents.tex
Almost all content is written.

@@
   We relied on reported verifications tasks, as recorded in bug reports, and interpreted the recorded data according to the documentation for the specific bug tracking system used.
-  As the partial results suggested that verification has no impact on software quality, we questioned the accuracy of the data about verification of bug fixes, ad thus decided to investigate how verification is actually performed and reported on specific software projects.
+  As the partial results suggested that verification has no impact on software quality, we questioned the accuracy of the data about verification of bug fixes, and thus decided to investigate how verification is actually performed and reported on specific software projects.
 
@@
 If it passes the quality requirements, it is CLOSED when the next release of the software is released.
-If it does not pass the quality requirements or if the solution only partially fixes the problem, the bug is REOPENed.
+If it does not pass the quality requirements or if the solution only partially fixes the problem, the bug is REOPENED.
 
@@
 
+\TODO bugs with status VERIFIED are used to write the release notes / change log.
+
 In order to understand the bug fixing process of a software team, however, it should be possible to know what activities are performed upon change of status. 
-For example, what steps are made before changing the status of a bug to RESOLVED (FIXED)?
+%For example, what steps are made before changing the status of a bug to RESOLVED (FIXED)?
 Is the patch applied to the version control system?
@@
 The documentation for Bugzilla does not prescribe specific activities for marking a bug as RESOLVED, relegating instead this decision to each software project.
-
-Similarly, what do developers do before marking a bug as VERIFIED?
+%
+%Similarly, w
+What do developers do before marking a bug as VERIFIED?
 Bugzilla documentation states that, when a bug is VERIFIED, it means that ``QA [quality assurance team] has looked at the bug and the resolution and agrees that the appropriate resolution has been taken''\footnote{\url{https://landfill.bugzilla.org/bugzilla-3.6-branch/page.cgi?id=fields.html}}.
@@
 Also, one cannot rely on the information provided by the bug tracking system or the developer documentation for a project, since it can be too generic to be useful.
-Such information can only be assessed by taking a close look on actual bug reports.
+Such information can only be assessed by taking a closer look on actual bug reports.
 
@@
 
-  \TODO the method has changed. Please ignore this section.
-
-In order to answer our exploratory research questions, we followed the following method:
+	In order to answer the research questions---when and how bug fixes are verified, and who verifies them---, a three-part method was used:
 
 	\begin{enumerate}
-	\item \textbf{Data extraction}: we have obtained publicly available raw data from the Bugzilla repositories of two popular integrated development environments, Eclipse and Bugzilla.
+		\item \textbf{Data extraction}: we have obtained publicly available raw data from the Bugzilla repositories of two popular integrated development environments, Eclipse and NetBeans.
 
-	\item \textbf{Data sampling}: we have chosen two subprojects. After that, we sampled bugs with interesting characteristics.
+		\item \textbf{Data sampling}: for each project, two representative subprojects were chosen for analysis.
 
-	\item \textbf{Data analysis}: we have read the full bug reported for the sampled bugs. Also, we have performed quantitative analysis on all bugs.
+		% we have chosen two subprojects. After that, we sampled bugs with interesting characteristics.
 
-	\item ...
+		% \item \textbf{Informal analysis}: some bug reports were fully read, so insights could be gained from them.
+
+		\item \textbf{Data analysis}: for each research question, a distinct analysis was required. The analyses will be further described.
 	\end{enumerate}
 
-\begin{subsection}{Available Data}
+\begin{subsection}{Data Extraction}
 
-	In order to perform the desired analysis, we needed access to all data recorded by Bugzilla for a specific project.
+	In order to perform the desired analyses, we needed access to the data recorded by Bugzilla for a specific project, including status changes and comments.
 We have found such data for two projects---Eclipse and NetBeans---from the domain of integrated development environments.
 The data were made available as part of the Mining Software Repositories 2011 Challenge\footnote{\url{http://www.msrconf.org/msr-challenge.html}} in the form of MySQL database dumps.
-The files contain all tables from the respective databases, except for the tables containing information about developers.
+The files contain all data from the respective databases, except for developer profiles, for privacy reasons.
 
@@
 It was licensed as open source in November, 2001.
-In 2004, IBM created the Eclipse Foundation in order to reduce the perception of Eclipse as a IBM-controlled project.
+The available data set contains 316,911 bug reports for its 155 subprojects, from October, 2001 to June, 2010.
 
-	NetBeans\footnote{\url{http://netbeans.org/about/history.html}} started  as a student project in 1996. It was bought by Sun Microsystems in October, 1999, and it was open sourced in March, 2000.
-
-	Both Eclipse and NetBeans are composed of many subprojects.
-The data comprises all bug reports from all Eclipse subprojects from October, 2001 to June, 2010, and from all NetBeans subprojects from June, 1998 to June, 2010.
-
-	There are 316,911 bug reports for 155 subprojects in Eclipse, and 185,578 bug reports for 39 subprojects in NetBeans.
+	NetBeans\footnote{\url{http://netbeans.org/about/history.html}} started  as a student project in 1996.
+It was bought by Sun Microsystems in October, 1999, and it was open sourced in March, 2000.
+The data set contains 185,578 bug reports for its 39 subprojects, from June, 1998 to June, 2010.
 
@@
 
-\begin{subsection}{Sampling of Bug Reports}
+\begin{subsection}{Data Sampling}
 
-	In order to obtain insight about the research questions, we read a sample of bug reports from both projects.
+	Four subprojects were chosen for further analysis: Eclipse/Platform, Eclipse/EMF, Netbeans/Platform, and Netbeans/VersionControl.
+The Platform subprojects are the main subprojects for the respective IDEs, so they are both important and representative of each projects' philosophy.
+
+	The other two subprojects were chosen at random, restricted to subprojects in which the proportion of verified bugs was greater than the proportion observed in the respective Platform subprojects.
+This is to avoid selecting projects in which bugs are seldom marked as VERIFIED.
+The proportion of VERIFIED bugs per project is as such: Eclipse/Platform: 16.0\%; Eclipse/EMF: 48.4\%; NetBeans/Platform: 21.4\%; NetBeans/VersionControl: 29.7\%.
+
+	% NetBeans/VersionControl: 0.29694479
+	% NetBeans/Platform: 0.21423101
+	% Eclipse/EMF: 0.4835361489
+	% Eclipse/Platform: 0.1595397738
+
+	% In order to obtain insight about the research questions, we read a sample of bug reports from both projects.
 For each project, two subprojects were chosen randomly among the projects with most verified bugs, resulting in the following subprojects: Eclipse/Platform, Eclipse/EMF, Netbeans/platform, Netbeans/versioncontrol.
 We have sampled 20 bug reports for each subproject, totalizing 80 bug reports.
+	% 
+	% \TODO this sampling was made just for the preliminary, qualitative analysis. Maybe we should describe preliminary analysis as an unstructured inquiry (instead of detailing the methods for it).
 
-	\TODO this sampling was made just for the preliminary, qualitative analysis. Maybe we should describe preliminary analysis as an unstructured inquiry (instead of detailing the methods for it).
+\end{subsection}
+
+\begin{subsection}{Analysis: How Are Bugs Verified?}
+
+	In order to discover the verification techniques used by the subprojects, we have looked at the comments written by developers when they mark a bug as VERIFIED (meaning that the fix was accepted) or REOPENED (meaning that the fix was rejected).
+First, an informal analysis was performed, by reading a sample of such messages, looking for textual patterns that indicate that a particular technique was used.
+Then, based on the informal analysis and on previous knowledge, we have built a regular expression for each technique, that should match messages in which there is evidence that that particular technique was used.
+The regular expressions were refined by looking at the results matched by them and modifying them to avoid false positives.
+Finally, we used the regular expressions to find relevant messages.
+
+	In this paper, we just list some substrings that the regular expression matches. The actual regular expressions are available at \url{http://...}.
+
+	\begin{itemize}
+		\item automated testing: testcase, test case, unit test, automated test, test result, run the test, test run;
+		\item source code inspection: inspecting, inspection, look at the code, looking at the source;
+		\item ad hoc testing: verified by ...ing;
+		\item automated static analysis: findbugs, checkstyle, static code, static analys.
+		\item formal testing: formal test, format verif
+	\end{itemize}
+
+	The classification script is smart enough to ignore negative assertions, such as ``did not run the test''.
+
+	The regular expression for ad hoc testing is the most restrictive of them, because ad hoc testing was found to be reported in many distinct ways (when it is reported).
+We expect to find fewer examples of ad hoc testing than it should be found.
+
+\end{subsection}
+
+\begin{subsection}{Analysis: When Are Bugs Verified?}
+
+	In order to determine if there is a well-defined verification phase for the subprojects, we have selected all reported verifications (i.e., status changes to VERIFIED) over the lifetime of each subproject.
+Then, we have plotted, for each day in the interval, the accumulated number of verifications reported since the beginning.
+The curve is monotonically increasing, with steeper climbs representing periods of intense verification activity.
+
+	Also, we have obtained the release dates for multiple versions of Eclipse and NetBeans.
+The information was obtained in the respective websites.
+In cases in which older information was not available, archived versions of the web pages were accessed via the website \url{www.archive.org}.
+
+	If a subproject presents a well-defined verification phase, it is expected that the verification activity is more intense a few days before a release.
+Such pattern can be identified by visual inspection of the graph, by looking for steeper climbs in the verification curve preceding the release dates.
+
+\end{subsection}
+
+\begin{subsection}{Analysis: Who Verifies Bugs?}
+
+	In order to determine whether there is a quality assurance (QA) team, we have counted how many times each developer has marked a bug as FIXED or VERIFIED.
+We considered that a developer is part of a QA team if s/he verified at least 10 times more than s/he fixed bugs.
+Also, we have computed the proportion of verifications for the subproject that was performed by the discovered QA team.
+It is expected that, if the discovered set of developer is actually a QA team, they should be responsible for the majority of the verifications.
 
@@
 
-\begin{subsection}{Preliminary Analysis}
-	Some insights were gained in this step, partially answering the who, when, and how questions:
+	Each of the three analysis was performed on Eclipse/Platform, Eclipse/EMF, NetBeans/Platform, and NetBeans/VersionControl. The results are presented.
 
-	\begin{itemize}
-		\item [how] In some projects, the status VERIFIED does not correspond to software verification.
+% \begin{subsection}{Preliminary Analysis}
+% 	Some insights were gained in this step, partially answering the who, when, and how questions:
+% 	
+% 	\begin{itemize}
+% 		\item [how] (OK) In some projects, the status VERIFIED does not correspond to software verification.
 In Eclipse/EMF, for example, it means that the fix is available in a build that is promoted on the project website\footnote{\url{http://wiki.eclipse.org/Modeling_PMC_Meeting,_2007-10-16} (between releases 3.3.1 and 3.3.1.1)}.
@@
 :-)''
-		\item [(who)] In the NetBeans subprojects that were analyzed, it was common that the reporter also verified the fix.
+% 		\item [(who)] (NOT RELEVANT) In the NetBeans subprojects that were analyzed, it was common that the reporter also verified the fix.
 In some bugs, this was necessary since the developer was not able to reproduce the bug.
-		\item [(who, when)] Often, the fixer verifies the bug just after marking it as fixed.
-		\item [(when)] Sometimes, several old bugs are marked as VERIFIED at the same time.
-		\item [(how?)] In both Eclipse and NetBeans, it is common that the fixer requests someone else to verify the fix (``Please verify, Tom'').
-		\item [(how)] In Eclipse/EMF, bug #249436, one developer rewrites the patch submitted by other developer.
-		\item [(how)] In Eclipse/EMF, bug #269789, a test case is used to assess the proposed fix.
-		\item [(how)] Eclipse/Platform, bug #210533: ``Verified by code inspection''.
-	\end{itemize}
+% 		\item [(who, when)] Often, the fixer verifies the bug just after marking it as fixed.
+% 		\item [(when)] (OK) Sometimes, several old bugs are marked as VERIFIED at the same time.
+% 		\item [(how?)] (OK) In both Eclipse and NetBeans, it is common that the fixer requests someone else to verify the fix (``Please verify, Tom'').
+% 		\item [(how)] (OK) In Eclipse/EMF, bug #249436, one developer rewrites the patch submitted by other developer.
+% 		\item [(how)] (OK) In Eclipse/EMF, bug #269789, a test case is used to assess the proposed fix.
+% 		\item [(how)] (OK) Eclipse/Platform, bug #210533: ``Verified by code inspection''.
+% 	\end{itemize}
+% 	
+% 	\TODO: ``Verified in/with [build]'', ``Verified by/through [verification technique]''.
+% 	
+% \end{subsection}
 
-	\TODO: ``Verified in/with [build]'', ``Verified by/through [verification technique]''.
+\begin{subsection}{How Are Bugs Verified?}
+
+  % To identify the verification techniques used by the projects, we looked at messages written by developers when they mark a bug as VERIFIED (meaning that the bug fix is appropriate) or REOPENED (meaning that some problem was found in the proposed fix).
+First, we have read some messages, non-systematically, to get some idea of what is said in theses messages and how it is said.
+
+  Most verification messages do not provide any clue about the verification technique used.
+Many messages just state that the bug fix was verified, optionally informing the build used to verify the fix.
+
+  In Eclipse/Platform, though, a significant number of verification messages include the technique used, e.g., ``verified by code inspection'', or ``verified with test case''.
+Based on such observations, we have built a search term for inspections and another one for unit tests.
+Then, we have classified the technique described in each verification message using the search term.
+
+  In Eclipse/Platform, there are 13,483 verifications, each one with a message.
+By the search criteria, 494 were identified as code inspections, 118 were identified as tests and 74 were identified as ad hoc.
+
+	Evidence of both code inspection and automated testing was found in all projects, although the evidence was weaker.
+
+  We also noted that, in Eclipse/Platform, often the fixer requests someone else to verify the fix, by asking ``please verify, [developer name]''.
+If the bug is reopened, then the fixer and the verifier exchange roles.
+That is considered good practice.
+
+  It is difficult to identify ad hoc cases, because they are so pervasive that they are not reported, and also they are reported in many different ways.
+
+	%We analyzed cases in which the bug is fixed, the fixer asks someone else to verify the fix, and then the bug is reopened.
+
+% \begin{itemize}
+% 	\item By creating or running unit tests.
+% 	\item By creating throw-away or configuration code that forces the occurrence of the error.
+% 	\item By running the software in a determined way (ad hoc). Often indicated by verbs.
+% 	\item By inspecting code.
+% \end{itemize}
+
+	Conclusion: it's hard to determine the technique used by just looking at the data. Eclipse/Platform documents it better, sometimes.
 
@@
 
-\begin{subsection}{RQ1: How Are Bugs Verified?}
-
-	We analyzed cases in which the bug is fixed, the fixer asks someone else to verify the fix, and then the bug is reopened.
-
-	Findings: ...
-
-\end{subsection}
-
-\begin{subsection}{RQ2: When Are Bugs Verified?}
+\begin{subsection}{When Are Bugs Verified?}
 
@@
 
+\begin{figure}
+\centering
+\mbox{
+\subfigure{\includegraphics[width=0.25\textwidth]{milestones}}%\quad
+\subfigure{\includegraphics[width=0.25\textwidth]{cliffs} }}
+\caption{Number of verifications over time.} \label{fig12}
+\end{figure}
+
+% \begin{figure}[htbp]
+% 	\centering
+% 		\includegraphics[scale=1]{file}
+% 	\caption{caption}
+% 	\label{fig:label}
+% \end{figure}
+
 	We found evidence of a verification phase for Eclipse/Platform.
 In this project, verifications appear to be performed in bursts, with periods of intense activity separated by about 6 to 7 weeks of low or no activity.
-By taking a closer look in the period between the releases 3.5 and 3.6, it is clear that the periods of intense verification activity occur just before the milestones leading to 3.6.
+By taking a closer look in the period between the releases 3.5 and 3.6 (Figure XXX), it is clear that the periods of intense verification activity occur just before the milestones leading to 3.6.
 
@@
 
-\begin{subsection}{RQ3: Who Verifies Bugs?}
+\begin{subsection}{Who Verifies Bugs?}
 
-	We counted the number of times each developer verified or fixed a bug.
+	Table \ref{tab:qa} presents, for each subproject, the number of developers attributed to a QA team, and the proportion of bug verifications they account for.
+	\newcommand\MyHead[2]{%
+	  \multicolumn{1}{l}{\parbox{#1}{\centering #2}}
+	}
+
+	\begin{table}
+		\begin{center}
+		\begin{tabular}{l|r|r}
+			% \toprule
+			\textbf{\centering Project} & 
+			\textbf{\centering QA team size} & 
+			\MyHead{2.5cm}{\textbf{\% of verifications by QA team}} \\
+			\hline
+			Eclipse/Platform        & 4 (2.4\%)  &  1.1\% \\
+			Eclipse/EMF             & 0 (0.0\%)  &  0.0\% \\
+			NetBeans/Platform       & 5 (21.7\%) & 93.3\% \\
+			NetBeans/VersionControl & 5 (20.8\%) & 93.2\% \\
+			% \bottomrule
+		\end{tabular}
+		\end{center}
+		\caption{Discovered QA team for all four subprojects.}
+		\label{tab:qa}
+	\end{table}
+
+	% We counted the number of times each developer verified or fixed a bug.
 We consider that, if a developer is primarily concerned with quality assurance, then s/he verifies much more bugs that s/he fixes.
+	% 
+	% We defined ratio as ...
+	% 
+	% The verified/fixed ratio for the four projects is shown on Table XXX.
 
-	We defined ratio as ...
-
-	The verified/fixed ratio for the four projects is shown on Table XXX.
-
-\begin{verbatim}
-*** eclipse-platform 
-Verifiers: 4 (2.4%)
-  1.1% of verifications
-Fixers: 30 (17.9%)
-Total developers: 168
-
-*** eclipse-emf 
-Verifiers: 0 (0.0%)
-  0.0% of verifications
-Fixers: 4 (30.8%)
-Total developers: 13
-
-*** netbeans-platform 
-Verifiers: 5 (21.7%)
-  93.3% of verifications
-Fixers: 6 (26.1%)
-Total developers: 23
-
-*** netbeans-versioncontrol 
-Verifiers: 5 (20.8%)
-  93.2% of verifications
-Fixers: 6 (25.0%)
-Total developers: 24 
-\end{verbatim}
+% \begin{verbatim}
+% *** eclipse-platform 
+% Verifiers: 4 (2.4%)
+%   1.1% of verifications
+% Fixers: 30 (17.9%)
+% Total developers: 168 
+% 
+% *** eclipse-emf 
+% Verifiers: 0 (0.0%)
+%   0.0% of verifications
+% Fixers: 4 (30.8%)
+% Total developers: 13 
+% 
+% *** netbeans-platform 
+% Verifiers: 5 (21.7%)
+%   93.3% of verifications
+% Fixers: 6 (26.1%)
+% Total developers: 23 
+% 
+% *** netbeans-versioncontrol 
+% Verifiers: 5 (20.8%)
+%   93.2% of verifications
+% Fixers: 6 (25.0%)
+% Total developers: 24 
+% \end{verbatim}
 
@@
 In Eclipse/EMF no developer focus specifically on verification.
-In Eclipse/Platform, only 2\% of the developers focus on verification tasks, however, they contribute only to 1\% of the verifications.
+In Eclipse/Platform, 2.4\% of the developers focus on verification tasks, contributing only to 1.1\% of the verifications.
 
@@
   The VERIFIED status is used in a variety of distinct ways, not always consistent with the traditional definition of software verification ...
+
+	Summary of the findings:
+	\begin{itemize}
+		\item It is difficult to detect the verification technique used in each reported verification. 
+		\item Beware of cliff walls!
+		\item Know your project. 
+	\end{itemize}
 

### 2012-02-04 contents.tex
Rewriting abstract and revising paper.

@@
 
+\newcommand{\UNCONFIRMED}{\texttt{UNCONFIRMED}}
+\newcommand{\NEW}{\texttt{NEW}}
+\newcommand{\ASSIGNED}{\texttt{ASSIGNED}}
+\newcommand{\VERIFIED}{\texttt{VERIFIED}}
+\newcommand{\RESOLVED}{\texttt{RESOLVED}}
+\newcommand{\CLOSED}{\texttt{CLOSED}}
+\newcommand{\REOPENED}{\texttt{REOPENED}}
+\newcommand{\FIXED}{\texttt{FIXED}}
+
 \begin{section}{Introduction}
@@
 Bug repositories (or bug tracking systems) have for a long time been used in software projects to support coordination among stakeholders.
-Such systems record discussion and progress of software evolution activities, such as corrective, perfective, and ...
-changes \cite{Antoniol2008}.
-Hence, bug repositories are an opportunity to researchers who intend to investigate issues related to the quality of the product and of the process of software development team.
+Such systems record discussion and progress of software evolution activities, such as bug fixing and software verification.
+Hence, bug repositories are an opportunity to researchers who intend to investigate issues related to the quality of both the product and the process of a software development team.
 %, or even issues regarding the nature of software development itself.
 
-The information contained in bug repositories has been used in order to predict fault-proneness, to unveil developers' roles from data, to characterize transfer of work in software projects, and so on.
+% The information contained in bug repositories has been used in order to predict fault-proneness, to unveil developers' roles from data, to characterize transfer of work in software projects, and so on.
 Also, it has been used to investigate beliefs in software engineering related to software quality, such as the impact of developer turnover and of collective ownership on the occurrence of bugs.
@@
 However, mining bug repositories has its own risks.
-Previous research has identified problems of missing data (e.g., rationale, traceability links between reported bug fixes and source code changes) and inaccurate data (e.g., misclassification of bugs) \cite{Aranda2009,Bird2009,Nguyen2010,Ayari2007}.
+Previous research has identified problems of missing data (e.g., rationale, traceability links between reported bug fixes and source code changes) and inaccurate data (e.g., misclassification of bugs) \cite{Aranda2009}.
+%,Bird2009,Nguyen2010,Ayari2007}.
 
-  In previous research [cite report], we tried to assess the impact of independent verification of bug fixes on software quality, by mining data from bug repositories. 
+  In previous research \cite{Souza2011}, we tried to assess the impact of independent verification of bug fixes on software quality, by mining data from bug repositories. 
   We relied on reported verifications tasks, as recorded in bug reports, and interpreted the recorded data according to the documentation for the specific bug tracking system used.
-  As the partial results suggested that verification has no impact on software quality, we questioned the accuracy of the data about verification of bug fixes, and thus decided to investigate how verification is actually performed and reported on specific software projects.
+  As the partial results suggested that verification has no impact on software quality, we questioned the accuracy of the data about verification of bug fixes, and thus decided to investigate how verification is actually performed and reported on the projects that were analyzed, Eclipse and NetBeans.
 
-	In particular, we are interested in answer the following exploratory research questions for a limited set of projects:
+	Hence, in this paper, we investigate the following exploratory research questions regarding the software verification process in Eclipse and NetBeans:
 
@@
 	\end{itemize}
+
+	The next section contains some background information about bug tracking systems and software verification.
+In Section \ref{sec:methods}, the data and methods used to investigate the research questions are presented.
+Then, in Section \ref{sec:results}, the results are exposed and discussed.
+Finally, Section \ref{sec:conclusion} presents some concluding remarks.
 
@@
 
-\begin{section}{Background}
+\begin{section}{Background} \label{sec:background}
 % Bug workflow. Bugzilla. \cite{Bird2009}
@@
 
-In the simplest cases, a bug is created and receive the status \texttt{UNCONFIRMED} (if it was created by a regular user) or \texttt{NEW} (if it was created by a developer).
-Next, it is ASSIGNED to a developer, and then it is RESOLVED, possibly by fixing it with a patch on the source code.
-The solution is then VERIFIED by someone in the quality assurance team.
-If it passes the quality requirements, it is CLOSED when the next release of the software is released.
-If it does not pass the quality requirements or if the solution only partially fixes the problem, the bug is REOPENED.
+In the simplest cases, a bug is created and receive the status \UNCONFIRMED\ (if it was created by a regular user) or \NEW\ (if it was created by a developer).
+Next, it is \ASSIGNED\ to a developer, and then it is \RESOLVED, possibly by fixing it with a patch on the source code.
+The solution is then \VERIFIED\ by someone in the quality assurance team.
+If it passes the quality requirements, it is \CLOSED\ when the next release of the software is released.
+If it does not pass the quality requirements or if the solution only partially fixes the problem, the bug is \REOPENED.
 
-\TODO Roles: reporter, fixer, verifier, reopener, closer.
+% \TODO Roles: reporter, fixer, verifier, reopener, closer.
 
-\TODO bugs with status VERIFIED are used to write the release notes / change log.
+% \TODO bugs with status \VERIFIED\ are used to write the release notes / change log.
 
@@
 %Similarly, w
-What do developers do before marking a bug as VERIFIED?
-Bugzilla documentation states that, when a bug is VERIFIED, it means that ``QA [quality assurance team] has looked at the bug and the resolution and agrees that the appropriate resolution has been taken''\footnote{\url{https://landfill.bugzilla.org/bugzilla-3.6-branch/page.cgi?id=fields.html}}.
+What do developers do before marking a bug as \VERIFIED?
+Bugzilla documentation states that, when a bug is \VERIFIED\, it means that ``QA [quality assurance team] has looked at the bug and the resolution and agrees that the appropriate resolution has been taken''\footnote{\url{https://landfill.bugzilla.org/bugzilla-3.6-branch/page.cgi?id=fields.html}}.
 Again, the definition is broad, although it assumes the existence of a QA team.
@@
 
-Software verification techniques are classified in static and dynamic [cite Sommervile].
+Software verification techniques are classified in static and dynamic \cite{Sommerville1995}.
 Static techniques include source code inspection, automated static analysis, and formal verification.
@@
 
-While mining bug repositories, one cannot assume that the VERIFIED status comprises all forms of software verification.
+While mining bug repositories, one cannot assume that the \VERIFIED\ status comprises all forms of software verification.
 Also, one cannot rely on the information provided by the bug tracking system or the developer documentation for a project, since it can be too generic to be useful.
@@
 
-\begin{section}{Method}
+\begin{section}{Method} \label{sec:method}
 
@@
 The data were made available as part of the Mining Software Repositories 2011 Challenge\footnote{\url{http://www.msrconf.org/msr-challenge.html}} in the form of MySQL database dumps.
-The files contain all data from the respective databases, except for developer profiles, for privacy reasons.
+The files contain all data from the respective databases, except for developer profiles, that were omitted for privacy reasons.
 
@@
 	The other two subprojects were chosen at random, restricted to subprojects in which the proportion of verified bugs was greater than the proportion observed in the respective Platform subprojects.
-This is to avoid selecting projects in which bugs are seldom marked as VERIFIED.
-The proportion of VERIFIED bugs per project is as such: Eclipse/Platform: 16.0\%; Eclipse/EMF: 48.4\%; NetBeans/Platform: 21.4\%; NetBeans/VersionControl: 29.7\%.
+The reason is to avoid selecting projects in which bugs are seldom marked as \VERIFIED.
+The following proportions of \VERIFIED\ bugs per project were observed: Eclipse/Platform: 16.0\%; Eclipse/EMF: 48.4\%; NetBeans/Platform: 21.4\%; NetBeans/VersionControl: 29.7\%.
 
@@
 
-	In order to discover the verification techniques used by the subprojects, we have looked at the comments written by developers when they mark a bug as VERIFIED (meaning that the fix was accepted) or REOPENED (meaning that the fix was rejected).
-First, an informal analysis was performed, by reading a sample of such messages, looking for textual patterns that indicate that a particular technique was used.
-Then, based on the informal analysis and on previous knowledge, we have built a regular expression for each technique, that should match messages in which there is evidence that that particular technique was used.
+	In order to discover the verification techniques used by the subprojects, we have looked at the comments written by developers when they mark a bug as \VERIFIED\ (meaning that the fix was accepted) or \REOPENED\ (meaning that the fix was rejected).
+First, an informal analysis was performed, by reading a sample of such comments, looking for textual patterns that indicate that a particular technique was used.
+Then, based on the informal analysis and on previous knowledge, a regular expression was built for each technique, that should match comments in which there is evidence that that particular technique was used.
 The regular expressions were refined by looking at the results matched by them and modifying them to avoid false positives.
-Finally, we used the regular expressions to find relevant messages.
 
-	In this paper, we just list some substrings that the regular expression matches. The actual regular expressions are available at \url{http://...}.
+	The regular expressions can be downloaded at \url{http://...}. For the sake of simplicity, the following list describes only some substrings matched by the regular expressions for each technique:
 
@@
 		\item source code inspection: inspecting, inspection, look at the code, looking at the source;
-		\item ad hoc testing: verified by ...ing;
-		\item automated static analysis: findbugs, checkstyle, static code, static analys.
-		\item formal testing: formal test, format verif
+		\item ad hoc testing: verified by running, verified by [verb]ing;
+		\item automated static analysis: findbugs, checkstyle, static code, static analys;
+		\item formal testing: formal test, format verif.
 	\end{itemize}
 
-	The classification script is smart enough to ignore negative assertions, such as ``did not run the test''.
+	% The classification script is smart enough to ignore negative assertions, such as ``did not run the test''.
 
-	The regular expression for ad hoc testing is the most restrictive of them, because ad hoc testing was found to be reported in many distinct ways (when it is reported).
-We expect to find fewer examples of ad hoc testing than it should be found.
+	During the informal analysis, it was observed that ad hoc testing is reported in many distinct ways that are difficult to capture by means of regular expressions.
+Because of this, the regular expression for is expected to match only a few comments.
+
+	% The regular expression for ad hoc testing is the most restrictive of them, because ad hoc testing was found to be reported in many distinct ways (when it is reported).
+%We expect to find fewer examples of ad hoc testing than it should be found.
 
@@
 
-	In order to determine if there is a well-defined verification phase for the subprojects, we have selected all reported verifications (i.e., status changes to VERIFIED) over the lifetime of each subproject.
-Then, we have plotted, for each day in the interval, the accumulated number of verifications reported since the beginning.
-The curve is monotonically increasing, with steeper climbs representing periods of intense verification activity.
+	In order to determine if there is a well-defined verification phase for the subprojects, we have selected all reported verifications (i.e., status changes to \VERIFIED) over the lifetime of each subproject.
+Then, we have plotted, for each day in the interval, the accumulated number of verifications reported since the first day available in the data.
+The curve is monotonically increasing, with steeper ascents representing periods of intense verification activity.
 
@@
 	If a subproject presents a well-defined verification phase, it is expected that the verification activity is more intense a few days before a release.
-Such pattern can be identified by visual inspection of the graph, by looking for steeper climbs in the verification curve preceding the release dates.
+Such pattern can be identified by visual inspection of the graph, by looking for steeper ascents in the verification curve preceding the release dates.
 
@@
 
-	In order to determine whether there is a quality assurance (QA) team, we have counted how many times each developer has marked a bug as FIXED or VERIFIED.
+	In order to determine whether there is a quality assurance (QA) team, we have counted how many times each developer has marked a bug as \FIXED\ or \VERIFIED.
 We considered that a developer is part of a QA team if s/he verified at least 10 times more than s/he fixed bugs.
-Also, we have computed the proportion of verifications for the subproject that was performed by the discovered QA team.
-It is expected that, if the discovered set of developer is actually a QA team, they should be responsible for the majority of the verifications.
+Also, we have computed the proportion of verifications that was performed by the discovered QA team.
+It is expected that, if the discovered set of developers is actually a QA team, they should be responsible for the majority of the verifications.
 
@@
 
-\begin{section}{Results and Discussion}
+\begin{section}{Results and Discussion} \label{sec:results}
 
-	Each of the three analysis was performed on Eclipse/Platform, Eclipse/EMF, NetBeans/Platform, and NetBeans/VersionControl. The results are presented.
+	Each of the three analysis was performed on Eclipse/Platform, Eclipse/EMF, NetBeans/Platform, and NetBeans/VersionControl. The results are presented next.
 
@@
 
-  Most verification messages do not provide any clue about the verification technique used.
-Many messages just state that the bug fix was verified, optionally informing the build used to verify the fix.
+  Most verification comments do not provide any clue about the verification technique used.
+Many comments just state that the bug fix was verified, optionally informing the build used to verify the fix.
 
-  In Eclipse/Platform, though, a significant number of verification messages include the technique used, e.g., ``verified by code inspection'', or ``verified with test case''.
+	\TODO table
+
+  % In Eclipse/Platform, though, a significant number of verification messages include the technique used, e.g., ``verified by code inspection'', or ``verified with test case''.
 Based on such observations, we have built a search term for inspections and another one for unit tests.
@@
 
+	\textbf{Observations}.
   We also noted that, in Eclipse/Platform, often the fixer requests someone else to verify the fix, by asking ``please verify, [developer name]''.
@@
   It is difficult to identify ad hoc cases, because they are so pervasive that they are not reported, and also they are reported in many different ways.
+
+	\TODO Eclipse/EMF: VERIFIED = in a build available in the website.
 
@@
 
-	In order to answer whether there is a verification phase for the projects analyzed, we plotted the cumulative number of verifications along the lifetime of each project, as shown in Figure XXX, in red.
+	% In order to answer whether there is a verification phase for the projects analyzed, we plotted the cumulative number of verifications along the lifetime of each project, as shown in Figure XXX, in red.
 The vertical lines represent release dates for each version.
 As a basis for comparison, the number of bug fixes is also plotted, in black.
+
+	Figure \ref{fig:when} shows plots of total accumulated number verifications over time for Eclipse/Platform (left) and NetBeans/Platform (right). Releases are plotted as dashed vertical lines.
 
@@
 \subfigure{\includegraphics[width=0.25\textwidth]{cliffs} }}
-\caption{Number of verifications over time.} \label{fig12}
+\caption{Number of verifications over time.} \label{fig:when}
 \end{figure}
@@
 
-	We found evidence of a verification phase for Eclipse/Platform.
+	For Eclipse/Platform, the graph shows the period between releases 3.5 and 3.6, including milestone releases every 6 to 7 weeks.
+It is clear from the graph that the verification activity is intensified in a few days preceding a milestone, represented in the graph by steeper ascents before the vertical lines.
+This pattern is an indicator of a verification phase.
+No such pattern was found in the other subprojects by analyzing their graphs (not shown here for brevity).
+
+	The graph for NetBeans/Platform (Figure \ref{fig:when}, right side) shows the entire project history.
+Although there are steeper ascents, they are different because they do not precede release dates.
+Also, at a closer look, they represent thousands of verifications performed in a few minutes by the same developer, with the same comment.
+The same pattern was found in Eclipse/EMF (not shown).
+
+	Of course, no developer can verify so many fixes in so little time.
+The explanation, supported by the comments, is that such mass verifications represent some cleanup of the bug repository, by marking old bugs as \VERIFIED---with no verification being actually performed.
+Researchers should take extra care with mass verifications, as they contain a large amount of bugs and, thus, are likely to bias the results of analyses.
+
+	% 
+	% We found evidence of a verification phase for Eclipse/Platform.
 In this project, verifications appear to be performed in bursts, with periods of intense activity separated by about 6 to 7 weeks of low or no activity.
 By taking a closer look in the period between the releases 3.5 and 3.6 (Figure XXX), it is clear that the periods of intense verification activity occur just before the milestones leading to 3.6.
-
-	No such pattern was found for Netbeans' subprojects.
+	% 
+	% No such pattern was found for Netbeans' subprojects.
 
@@
 
-	The first thing that comes to attention are ``cliff walls'', points in time when a large amount of bugs are verified, represented by steep, almost vertical portions of the plotted line.
+	% The first thing that comes to attention are ``cliff walls'', points in time when a large amount of bugs are verified, represented by steep, almost vertical portions of the plotted line.
 They appear in Eclipse/EMF and Netbeans/Platform.
 In Netbeans/Platform, all verifications in the cliff wall preceding version 5.5 share the same message: ``Verification of old issues''.
-For Eclipse/EMF, the message is ``Move to verified as per bug 206558'', which refers to the task of changing bugs marked as FIXED to VERIFIED.
+For Eclipse/EMF, the message is ``Move to verified as per bug 206558'', which refers to the task of changing bugs marked as \FIXED\ to \VERIFIED.
+	% 
+	% Cliff walls represent a change in the development process, and the mass verification is needed in order to leverage useful features from Bugzilla (e.g., looking for verified bugs in order to aid the writing of release notes).
 
-	Cliff walls represent a change in the development process, and the mass verification is needed in order to leverage useful features from Bugzilla (e.g., looking for verified bugs in order to aid the writing of release notes).
-
-	Because of cliff walls, extra care must be taken before doing any analysis that relies on the VERIFIED status.
-In such situations, the VERIFIED status is applied blindly, with no software verification being actually performed.
+	% Because of cliff walls, extra care must be taken before doing any analysis that relies on the \VERIFIED\ status.
+In such situations, the \VERIFIED\ status is applied blindly, with no software verification being actually performed.
 Because cliff walls contain a large amount of bugs, they are likely to bias the results of analyses.
@@
 
-	Table \ref{tab:qa} presents, for each subproject, the number of developers attributed to a QA team, and the proportion of bug verifications they account for.
+	Table \ref{tab:qa} presents, for each subproject, the number of developers attributed to a QA team (i.e., developers who perform verifications at least 10 times more often than contribute bug fixes), and the proportion of bug verifications they account for.
 	\newcommand\MyHead[2]{%
@@
 
-	In NetBeans it is possible to infer the existence of a QA team, composed by approximately 20\% of the developers, that performs about 90\% of all verifications.
-
+	In NetBeans (both Platform and VersionControl) it is possible to infer the existence of a QA team, composed by approximately 20\% of the developers, that performs about 90\% of all verifications.
 In the Eclipse subprojects, there is no evidence of a dedicated QA team.
@@
 
-\begin{section}{Conclusion}
-  The VERIFIED status is used in a variety of distinct ways, not always consistent with the traditional definition of software verification ...
+\begin{section}{Conclusion} \label{sec:conclusion}
+  The \VERIFIED\ status is used in a variety of distinct ways, not always consistent with the traditional definition of software verification ...
 

### 2012-02-04 contents.tex
revision.

@@
 
-% \TODO why the VERIFIED status is so important?
-
 % In summary, for researchers who intend to use data from bug tracking systems, it is not sufficient to understand the data fields in a bug report the way they are documented in the system's documentation.
@@
 
-% \TODO Roles: reporter, fixer, verifier, reopener, closer.
+% Roles: reporter, fixer, verifier, reopener, closer.
 
-% \TODO bugs with status \VERIFIED\ are used to write the release notes / change log.
+% bugs with status \VERIFIED\ are used to write the release notes / change log.
 
@@
 
-\begin{section}{Method} \label{sec:method}
+\begin{section}{Method} \label{sec:methods}
 
@@
 	%
-	% \TODO this sampling was made just for the preliminary, qualitative analysis. Maybe we should describe preliminary analysis as an unstructured inquiry (instead of detailing the methods for it).
 
@@
 		\item source code inspection: inspecting, inspection, look at the code, looking at the source;
-		\item ad hoc testing: verified by running, verified by [verb]ing;
+		\item ad hoc testing: verified by running, verified by ...ing;
 		\item automated static analysis: findbugs, checkstyle, static code, static analys;
-		\item formal testing: formal test, format verif.
+		\item formal testing: formal test, formal verif.
 	\end{itemize}
@@
 % 	
-% 	\TODO: ``Verified in/with [build]'', ``Verified by/through [verification technique]''.
-% 	
 % \end{subsection}
@@
   Most verification comments do not provide any clue about the verification technique used.
-Many comments just state that the bug fix was verified, optionally informing the build used to verify the fix.
+Most often, developers just state that the bug fix was verified, sometimes informing the build used to verify the fix.
 
-	\TODO table
+	Table \ref{tab:how} shows, for each subproject, the number of comments associated with a bug being marked as \VERIFIED\ or \REOPENED\ that refers to a particular verification technique.
+No references to formal verification or static analysis were found; therefore, these techniques are not shown on the table.
+
+	\begin{table}
+		\begin{center}
+		\caption{Verification techniques for all four subprojects.}
+		\begin{tabular}{l|r|r|r}
+			% \toprule
+			\textbf{\centering Project} & 
+			\textbf{\centering Testing} & 
+			\textbf{\centering Inspection} & 
+			\textbf{\centering Ad Hoc} & 
+			\hline
+			Eclipse/Platform        & 260 (1.1\%) & 524 (2.2\%) & 74 (0.3\%) \\
+			Eclipse/EMF             & 27 (0.7\%) & 2 (0.1\%) & 0 (0.0\%) \\
+			NetBeans/Platform       & 94 (0.6\%) & 5 (0.0\%) & 0 (0.0\%) \\
+			NetBeans/VersionControl & 4 (0.1\%) & 1 (0.0\%) & 0 (0.0\%) 
+			% \bottomrule
+		\end{tabular}
+		\end{center}
+
+		\label{tab:how}
+	\end{table}
+
+	In Eclipse/Platform, less than 4\% of the comments referenced the verification technique that was used; for the other projects, less than 1\% of the comments contained such references.
+Such low proportion reveals how difficult it is to infer the verification technique that was applied using only data from bug repositories.
+This is probably tacit knowledge that can only be extracted by asking the developers themselves.
+
+	In all subprojects, comments suggest the use of automated testing and code inspection in the verification process, the former technique being more frequently referenced (except in Eclipse/Platform, in which code inspection is cited more often).
+Evidences of ad hoc testing were found only in Eclipse/Platform, because verification comments tend to be more structured in this project, but we believe that ad hoc testing is much more common.
+
+	While sorting through comments, we have found interesting information regarding the software verification process.
+In Eclipse/Platform, the developer who fixes a bug often someone else to verify the fix, by asking ``please verify, [developer name]''.
+If the bug is reopened, then the fixer and the verifier exchange roles.
+This behavior illustrates a structured bug fixing/verification process.
+
+	In Eclipse/EMF, we found that marking a bug as \VERIFIED\ does not mean that the bug fix was verified.
+Instead, it means that the fix was made available in a build of the software that is published in the subproject's website.
+This reinterpretation of the Bugzilla workflow was decided in a meeting\footnote{\url{http://wiki.eclipse.org/Modeling_PMC_Meeting,_2007-10-16}}.
 
@@
 
-  In Eclipse/Platform, there are 13,483 verifications, each one with a message.
+	%   In Eclipse/Platform, there are 13,483 verifications, each one with a message.
 By the search criteria, 494 were identified as code inspections, 118 were identified as tests and 74 were identified as ad hoc.
+	% 
+	% Evidence of both code inspection and automated testing was found in all projects, although the evidence was weaker.
 
-	Evidence of both code inspection and automated testing was found in all projects, although the evidence was weaker.
-
-	\textbf{Observations}.
-  We also noted that, in Eclipse/Platform, often the fixer requests someone else to verify the fix, by asking ``please verify, [developer name]''.
+	% \textbf{Observations}.
+	%   We also noted that, in Eclipse/Platform, often the fixer requests someone else to verify the fix, by asking ``please verify, [developer name]''.
 If the bug is reopened, then the fixer and the verifier exchange roles.
 That is considered good practice.
+	% 
+	%   It is difficult to identify ad hoc cases, because they are so pervasive that they are not reported, and also they are reported in many different ways.
 
-  It is difficult to identify ad hoc cases, because they are so pervasive that they are not reported, and also they are reported in many different ways.
-
-	\TODO Eclipse/EMF: VERIFIED = in a build available in the website.
+	% Eclipse/EMF: VERIFIED = in a build available in the website.
 
@@
 		\begin{center}
+
+		\caption{Discovered QA team for all four subprojects.}
 		\begin{tabular}{l|r|r}
@@
 		\end{center}
-		\caption{Discovered QA team for all four subprojects.}
+
 		\label{tab:qa}
@@
 \begin{section}{Conclusion} \label{sec:conclusion}
-  The \VERIFIED\ status is used in a variety of distinct ways, not always consistent with the traditional definition of software verification ...
+	By analyzing four subprojects (two from Eclipse, two from NetBeans), we have found, using only data from bug repositories, subprojects with and without QA teams, with and without a well-defined verification phase.
+We also have found evidence of the application of automated testing and source code inspection, although such evidence was less reliable.
+Also, we have found cases in which marking a bug as \VERIFIED\ does not imply that any kind of software verification was actually performed.
 
-	Summary of the findings:
-	\begin{itemize}
-		\item It is difficult to detect the verification technique used in each reported verification. 
-		\item Beware of cliff walls!
-		\item Know your project. 
-	\end{itemize}
+	With the knowledge obtained from this exploratory research, we aim to improve and extend our previous work on the impact of independent verification on software quality.
+We can investigate, for example, whether verification performed by QA team is better than verification performed by other developers.
 
-	With the knowledge obtained from this exploratory research, we can improve and extend our previous work on the impact of independent verification on software quality. Examples of refined questions:
+	Researchers should be aware that information about verification techniques are not common in bug repositories, and that reported verification does not always correspond to actual verification.
+Some exploration of the data is important to avoid such pitfalls.
 
-	\begin{itemize}
-		\item Does independent verification improve software quality when it is performed by developers from the QA team?
-		\item Does independent verification improve software quality when it is performed as an separate phase of the software release cycle?
-		\item Does independent verification improve software quality when a particular verification technique (e.g., inspections, unit testing) is used?
-	\end{itemize}
+	% \begin{itemize}
+	% 	\item Does independent verification improve software quality when it is performed by developers from the QA team?
+	% 	\item Does independent verification improve software quality when it is performed as an separate phase of the software release cycle?
+	% 	\item Does independent verification improve software quality when a particular verification technique (e.g., inspections, unit testing) is used?
+	% \end{itemize}
 

### 2012-02-04 contents.tex
Revising up to section 2 (inclusive).

@@
 
-Bug repositories (or bug tracking systems) have for a long time been used in software projects to support coordination among stakeholders.
-Such systems record discussion and progress of software evolution activities, such as bug fixing and software verification.
-Hence, bug repositories are an opportunity to researchers who intend to investigate issues related to the quality of both the product and the process of a software development team.
+Bug repositories have for a long time been used in software projects to support coordination among stakeholders.
+They record discussion and progress of software evolution activities, such as bug fixing and software verification.
+Hence, bug repositories are an opportunity for researchers who intend to investigate issues related to the quality of both the product and the process of a software development team.
 %, or even issues regarding the nature of software development itself.
@@
 		\item \textbf{How} is the verification performed: are there performed ad hoc tests, automated tests, code inspection?
-		\item \textbf{Who} performs the verification: is there a dedicated team for QA? 
+		\item \textbf{Who} performs the verification: is there a QA (quality assurance) team? 
 		\item \textbf{When} is the verification performed: is it performed just after the fix, or is there a verification phase?
@@
 
-Bug tracking systems allow users and developers of a software project to manage a list of bugs for the project.
-Usually, users and developers can report bugs, along with information such as steps to reproduce the bug, its severity and the operating system used.
+Bug tracking systems allow users and developers of a software project to manage a list of bugs for the project, along with information such as steps to reproduce the bug, its severity and the operating system used.
 Developers choose bugs to fix and can report on the progress of the bug fixing activities, ask for clarification, discuss causes for the bug etc.
 
-In this research, we focus on Bugzilla, an open source bug tracking system used by notable software projects such as Eclipse, Mozilla, Linux Kernel, Open Office, Apache, and companies such as NASA and Facebook\footnote{Complete list available at http://www.bugzilla.org/installation-list/.}.
+In this research, we focus on Bugzilla, an open source bug tracking system used by notable software projects such as Eclipse, Mozilla, Linux Kernel, NetBeans, Apache, and companies such as NASA and Facebook\footnote{Complete list available at http://www.bugzilla.org/installation-list/.}.
 The general concepts from Bugzilla should apply to most other bug tracking systems.
 
-One important feature of a bug that is recorded on bug tracking systems is its status.
-The status records the progress of the bug fixing activity, and, as such, provides data to software engineering studies.
+One important feature of a bug that is recorded on bug tracking systems is its status. The status records the progress of the bug fixing activity.
+%, and, as such, provides data to software engineering studies
 Figure \ref{fig:bugzilla} shows each status that can be recorded in Bugzilla, along with typical transitions between status values, i.e., the workflow.
@@
 Next, it is \ASSIGNED\ to a developer, and then it is \RESOLVED, possibly by fixing it with a patch on the source code.
-The solution is then \VERIFIED\ by someone in the quality assurance team.
-If it passes the quality requirements, it is \CLOSED\ when the next release of the software is released.
+The solution is then \VERIFIED\ by someone in the quality assurance team, if it is adequate, or otherwise it is \REOPENED.
+When a version of the software is released, all \VERIFIED\ bugs are \CLOSED.
+
+% If it passes the quality requirements, it is \CLOSED\ when the next release of the software is released.
 If it does not pass the quality requirements or if the solution only partially fixes the problem, the bug is \REOPENED.
@@
 
-In order to understand the bug fixing process of a software team, however, it should be possible to know what activities are performed upon change of status. 
+% In order to understand the bug fixing process of a software team, however, it should be possible to know what activities are performed upon change of status. 
 %For example, what steps are made before changing the status of a bug to RESOLVED (FIXED)?
@@
 %Similarly, w
-What do developers do before marking a bug as \VERIFIED?
-Bugzilla documentation states that, when a bug is \VERIFIED\, it means that ``QA [quality assurance team] has looked at the bug and the resolution and agrees that the appropriate resolution has been taken''\footnote{\url{https://landfill.bugzilla.org/bugzilla-3.6-branch/page.cgi?id=fields.html}}.
-Again, the definition is broad, although it assumes the existence of a QA team.
+%What do developers do before marking a bug as \VERIFIED? 
+Bugzilla documentation states that, when a bug is \VERIFIED, it means that ``QA [quality assurance team] has looked at the bug and the resolution and agrees that the appropriate resolution has been taken''\footnote{\url{https://landfill.bugzilla.org/bugzilla-3.6-branch/page.cgi?id=fields.html}}.
+It does not specify how developers should look at the resolution (e.g., by looking at the source code, or by running a patched version of the software).
 
-The developer documentation for both Eclipse and NetBeans are unclear about how the verification of a bug fix is performed.
+% Again, the definition is broad, although it assumes the existence of a QA team.
+
+% The developer documentation for both Eclipse and NetBeans are unclear about how the verification of a bug fix is performed.
 So, what forms of  software verification are they used in the verification process of these projects?
@@
 Software verification techniques are classified in static and dynamic \cite{Sommerville1995}.
-Static techniques include source code inspection, automated static analysis, and formal verification.
+Static techniques include \emph{source code inspection}, automated \emph{static analysis}, and \emph{formal verification}.
 Dynamic techniques, or testing, involve executing the software system under certain conditions and comparing its actual behavior with the intended behavior.
-Testing can be done in an improvised way (ad hoc testing), or can be compiled in a list of test cases, which can then be automated (e.g., as unit tests).
+Testing can be done in an improvised way (\emph{ad hoc testing}), or it can be structured in a list of test cases, leading to \emph{automated testing}.
 
-While mining bug repositories, one cannot assume that the \VERIFIED\ status comprises all forms of software verification.
+% While mining bug repositories, one cannot assume that the \VERIFIED\ status comprises all forms of software verification.
 Also, one cannot rely on the information provided by the bug tracking system or the developer documentation for a project, since it can be too generic to be useful.

### 2012-02-04 contents.tex
Version 1.

@@
 However, mining bug repositories has its own risks.
-Previous research has identified problems of missing data (e.g., rationale, traceability links between reported bug fixes and source code changes) and inaccurate data (e.g., misclassification of bugs) \cite{Aranda2009}.
+Previous research has identified problems of missing data (e.g., rationale, traceability links between reported bug fixes and source code changes) and inaccurate data (e.g., misclassification of bugs) \cite{Aranda2009,Bird2009}.
 %,Bird2009,Nguyen2010,Ayari2007}.
@@
 The data were made available as part of the Mining Software Repositories 2011 Challenge\footnote{\url{http://www.msrconf.org/msr-challenge.html}} in the form of MySQL database dumps.
-The files contain all data from the respective databases, except for developer profiles, that were omitted for privacy reasons.
+The files contain all data from the respective databases, except for developer profiles, omitted for privacy reasons.
 
@@
 	NetBeans\footnote{\url{http://netbeans.org/about/history.html}} started  as a student project in 1996.
-It was bought by Sun Microsystems in October, 1999, and it was open sourced in March, 2000.
+It was then bought by Sun Microsystems in October, 1999, and open sourced in March, 2000.
 The data set contains 185,578 bug reports for its 39 subprojects, from June, 1998 to June, 2010.
@@
 
-	Four subprojects were chosen for further analysis: Eclipse/Platform, Eclipse/EMF, Netbeans/Platform, and Netbeans/VersionControl.
+	Four subprojects were chosen for further analysis: Eclipse/Platform, Eclipse/EMF, NetBeans/Platform, and NetBeans/VersionControl.
 The Platform subprojects are the main subprojects for the respective IDEs, so they are both important and representative of each projects' philosophy.
@@
 	% In order to obtain insight about the research questions, we read a sample of bug reports from both projects.
-For each project, two subprojects were chosen randomly among the projects with most verified bugs, resulting in the following subprojects: Eclipse/Platform, Eclipse/EMF, Netbeans/platform, Netbeans/versioncontrol.
+For each project, two subprojects were chosen randomly among the projects with most verified bugs, resulting in the following subprojects: Eclipse/Platform, Eclipse/EMF, NetBeans/platform, NetBeans/versioncontrol.
 We have sampled 20 bug reports for each subproject, totalizing 80 bug reports.
@@
 First, an informal analysis was performed, by reading a sample of such comments, looking for textual patterns that indicate that a particular technique was used.
-Then, based on the informal analysis and on previous knowledge, a regular expression was built for each technique, that should match comments in which there is evidence that that particular technique was used.
+Then, based on the informal analysis and on previous knowledge, a regular expression was built for each technique, to match comments in which there is evidence that that particular technique was used.
 The regular expressions were refined by looking at the results matched by them and modifying them to avoid false positives.
 
-	The regular expressions can be downloaded at \url{http://...}. For the sake of simplicity, the following list describes only some substrings matched by the regular expressions for each technique:
+	The regular expressions are available at \url{http://...}. For the sake of simplicity, the following list describes only some substrings matched by the regular expressions for each technique:
 
@@
 	During the informal analysis, it was observed that ad hoc testing is reported in many distinct ways that are difficult to capture by means of regular expressions.
-Because of this, the regular expression for is expected to match only a few comments.
+Hence, its regular expression is expected to match only a few comments.
 
@@
 
-  Most verification comments do not provide any clue about the verification technique used.
-Most often, developers just state that the bug fix was verified, sometimes informing the build used to verify the fix.
-
 	Table \ref{tab:how} shows, for each subproject, the number of comments associated with a bug being marked as \VERIFIED\ or \REOPENED\ that refers to a particular verification technique.
@@
 			\textbf{\centering Inspection} & 
-			\textbf{\centering Ad Hoc} & 
+			\textbf{\centering Ad Hoc} \\
 			\hline
@@
 	While sorting through comments, we have found interesting information regarding the software verification process.
+Most verification comments do not provide any clue about the verification technique used.
+Most often, developers just state that the bug fix was verified, sometimes informing the build used to verify the fix.
+
 	In Eclipse/Platform, the developer who fixes a bug often someone else to verify the fix, by asking ``please verify, [developer name]''.
@@
 
-	Conclusion: it's hard to determine the technique used by just looking at the data. Eclipse/Platform documents it better, sometimes.
+	% Conclusion: it's hard to determine the technique used by just looking at the data. Eclipse/Platform documents it better, sometimes.
 
@@
 
-	Figure \ref{fig:when} shows plots of total accumulated number verifications over time for Eclipse/Platform (left) and NetBeans/Platform (right). Releases are plotted as dashed vertical lines.
+	Figure \ref{fig:when} shows plots of total accumulated number of verifications over time for Eclipse/Platform (left) and NetBeans/Platform (right). Releases are plotted as dashed vertical lines.
 
@@
 
-	For Eclipse/Platform, the graph shows the period between releases 3.5 and 3.6, including milestone releases every 6 to 7 weeks.
+	For Eclipse/Platform, the graph shows the period between releases 3.5 and 3.6, including milestone releases every 6 or 7 weeks.
 It is clear from the graph that the verification activity is intensified in a few days preceding a milestone, represented in the graph by steeper ascents before the vertical lines.
@@
 	% 
-	% No such pattern was found for Netbeans' subprojects.
+	% No such pattern was found for NetBeans' subprojects.
 
@@
 	% The first thing that comes to attention are ``cliff walls'', points in time when a large amount of bugs are verified, represented by steep, almost vertical portions of the plotted line.
-They appear in Eclipse/EMF and Netbeans/Platform.
-In Netbeans/Platform, all verifications in the cliff wall preceding version 5.5 share the same message: ``Verification of old issues''.
+They appear in Eclipse/EMF and NetBeans/Platform.
+In NetBeans/Platform, all verifications in the cliff wall preceding version 5.5 share the same message: ``Verification of old issues''.
 For Eclipse/EMF, the message is ``Move to verified as per bug 206558'', which refers to the task of changing bugs marked as \FIXED\ to \VERIFIED.
@@
 
-	In NetBeans (both Platform and VersionControl) it is possible to infer the existence of a QA team, composed by approximately 20\% of the developers, that performs about 90\% of all verifications.
+	In both NetBeans subprojects it is possible to infer the existence of a QA team, composed by approximately 20\% of the developers, that performs about 90\% of all verifications.
 In the Eclipse subprojects, there is no evidence of a dedicated QA team.
@@
 	By analyzing four subprojects (two from Eclipse, two from NetBeans), we have found, using only data from bug repositories, subprojects with and without QA teams, with and without a well-defined verification phase.
-We also have found evidence of the application of automated testing and source code inspection, although such evidence was less reliable.
+We also have found weaker evidence of the application of automated testing and source code inspection.
 Also, we have found cases in which marking a bug as \VERIFIED\ does not imply that any kind of software verification was actually performed.
@@
 	With the knowledge obtained from this exploratory research, we aim to improve and extend our previous work on the impact of independent verification on software quality.
-We can investigate, for example, whether verification performed by QA team is better than verification performed by other developers.
+We can investigate, for example, whether verification performed by QA team is more effective than verification performed by other developers.
 

### 2012-02-06 contents.tex
Applying suggestions by Christina.

@@
 However, mining bug repositories has its own risks.
-Previous research has identified problems of missing data (e.g., rationale, traceability links between reported bug fixes and source code changes) and inaccurate data (e.g., misclassification of bugs) \cite{Aranda2009,Bird2009}.
+Previous research has identified problems of missing data (e.g., rationale, traceability links between reported bug fixes and source code changes) and inaccurate data (e.g., misclassification of bugs) \cite{Aranda2009}.
 %,Bird2009,Nguyen2010,Ayari2007}.
@@
 % The developer documentation for both Eclipse and NetBeans are unclear about how the verification of a bug fix is performed.
-So, what forms of  software verification are they used in the verification process of these projects?
+So, what are the forms of software verification used in the verification process of these projects?
+
 % In some parts they mention running unit tests, and running a patched version of the software and try to reproduce the bug.
@@
 Dynamic techniques, or testing, involve executing the software system under certain conditions and comparing its actual behavior with the intended behavior.
-Testing can be done in an improvised way (\emph{ad hoc testing}), or it can be structured in a list of test cases, leading to \emph{automated testing}.
+Testing can be done in an improvised way (\emph{ad hoc testing}), or it can be structured as a list of test cases, leading to \emph{automated testing}.
 
@@
 Also, one cannot rely on the information provided by the bug tracking system or the developer documentation for a project, since it can be too generic to be useful.
-Such information can only be assessed by taking a closer look on actual bug reports.
+Such information can be assessed by taking a closer look on actual bug reports.
 
@@
 	\end{enumerate}
+
+	The experimental package is available at \url{https://sites.google.com/site/rodrigorgs2/msr2012}.
 
@@
 
-	The regular expressions are available at \url{http://...}. For the sake of simplicity, the following list describes only some substrings matched by the regular expressions for each technique:
+	The complete regular expressions are available in the experimental package.
+For the sake of simplicity, the following list describes only some substrings matched by the regular expressions for each technique:
 
@@
 		\item source code inspection: inspecting, inspection, look at the code, looking at the source;
-		\item ad hoc testing: verified by running, verified by ...ing;
+		\item ad hoc testing: verified by running, verified by [word ending in ``ing''];
 		\item automated static analysis: findbugs, checkstyle, static code, static analys;
@@
 
-	In order to determine whether there is a quality assurance (QA) team, we have counted how many times each developer has marked a bug as \FIXED\ or \VERIFIED.
-We considered that a developer is part of a QA team if s/he verified at least 10 times more than s/he fixed bugs.
+	In order to determine whether there is a team dedicated to quality assurance (QA), we have counted how many times each developer has marked a bug as \FIXED\ or \VERIFIED.
+We considered that a developer is part of a QA team if s/he verified at least 10 times (i.e., one order of magnitude) more than s/he fixed bugs.
 Also, we have computed the proportion of verifications that was performed by the discovered QA team.
@@
 	In all subprojects, comments suggest the use of automated testing and code inspection in the verification process, the former technique being more frequently referenced (except in Eclipse/Platform, in which code inspection is cited more often).
-Evidences of ad hoc testing were found only in Eclipse/Platform, because verification comments tend to be more structured in this project, but we believe that ad hoc testing is much more common.
+Evidences of ad hoc testing were found only in Eclipse/Platform, because verification comments tend to be more structured in this project, but we believe that ad hoc testing is much more common, given its unstructured nature.
 
@@
 
-	Researchers should be aware that information about verification techniques are not common in bug repositories, and that reported verification does not always correspond to actual verification.
+	Researchers should be aware that information about verification techniques may not be common in bug repositories, and that reported verification does not always correspond to actual verification.
 Some exploration of the data is important to avoid such pitfalls.

### 2012-02-07 contents.tex
Version submitted to MSR 2012.

@@
 
-	The experimental package is available at \url{https://sites.google.com/site/rodrigorgs2/msr2012}.
+	The experimental package is available at \\ \url{https://sites.google.com/site/rodrigorgs2/msr2012}
 
@@
 		\begin{center}
-		\caption{Verification techniques for all four subprojects.}
+		\caption{Verification techniques for all four subprojects.} \label{tab:how}
 		\begin{tabular}{l|r|r|r}
@@
 		\end{center}
-
-		\label{tab:how}
 	\end{table}
@@
 
-		\caption{Discovered QA team for all four subprojects.}
+		\caption{Discovered QA team for all four subprojects.} \label{tab:qa}
 		\begin{tabular}{l|r|r}
@@
 
-		\label{tab:qa}
 	\end{table}
@@
 We also have found weaker evidence of the application of automated testing and source code inspection.
-Also, we have found cases in which marking a bug as \VERIFIED\ does not imply that any kind of software verification was actually performed.
+Also, there were cases in which marking a bug as \VERIFIED\ did not imply that any kind of software verification was actually performed.
 

### 2012-03-28 contents.tex
Minor changes after conference review.

@@
 However, mining bug repositories has its own risks.
-Previous research has identified problems of missing data (e.g., rationale, traceability links between reported bug fixes and source code changes) and inaccurate data (e.g., misclassification of bugs) \cite{Aranda2009}.
+Previous research has identified problems of missing data (e.g., rationale, traceability links between reported bug fixes and source code changes), inaccurate data (e.g., misclassification of bugs) \cite{Aranda2009}, and biased data \cite{Bird2009}.
 %,Bird2009,Nguyen2010,Ayari2007}.
@@
 	\begin{itemize}
+		\item \textbf{When} is the verification performed: is it performed just after the fix, or is there a verification phase?
+		\item \textbf{Who} performs the verification: is there a QA (quality assurance) team? 		
 		\item \textbf{How} is the verification performed: are there performed ad hoc tests, automated tests, code inspection?
-		\item \textbf{Who} performs the verification: is there a QA (quality assurance) team? 
-		\item \textbf{When} is the verification performed: is it performed just after the fix, or is there a verification phase?
 		% \item How is the distribution of verification across developers?
@@
 We have found such data for two projects---Eclipse and NetBeans---from the domain of integrated development environments.
-The data were made available as part of the Mining Software Repositories 2011 Challenge\footnote{\url{http://www.msrconf.org/msr-challenge.html}} in the form of MySQL database dumps.
+The data was made available as part of the Mining Software Repositories 2011 Challenge\footnote{\url{http://2011.msrconf.org/msr-challenge.html}} in the form of MySQL database dumps.
 The files contain all data from the respective databases, except for developer profiles, omitted for privacy reasons.
@@
 	Also, we have obtained the release dates for multiple versions of Eclipse and NetBeans.
-The information was obtained in the respective websites.
+The information was obtained from the respective websites.
 In cases in which older information was not available, archived versions of the web pages were accessed via the website \url{www.archive.org}.
@@
 
-	In Eclipse/Platform, the developer who fixes a bug often someone else to verify the fix, by asking ``please verify, [developer name]''.
+	In Eclipse/Platform, the developer who fixes a bug often asks someone else to verify the fix, by asking ``please verify, [developer name]''.
 If the bug is reopened, then the fixer and the verifier exchange roles.
@@
 
-The first author is supported by FAPESB under grant BOL0119/2010.
+\small{This work is supported by FAPESB (grant BOL0119/2010) and CNPq/INES (grant 573964/2008-4).}

### 2012-03-28 contents.tex
Changing order of results.

@@
 
-\begin{subsection}{Analysis: How Are Bugs Verified?}
-
-	In order to discover the verification techniques used by the subprojects, we have looked at the comments written by developers when they mark a bug as \VERIFIED\ (meaning that the fix was accepted) or \REOPENED\ (meaning that the fix was rejected).
-First, an informal analysis was performed, by reading a sample of such comments, looking for textual patterns that indicate that a particular technique was used.
-Then, based on the informal analysis and on previous knowledge, a regular expression was built for each technique, to match comments in which there is evidence that that particular technique was used.
-The regular expressions were refined by looking at the results matched by them and modifying them to avoid false positives.
-
-	The complete regular expressions are available in the experimental package.
-For the sake of simplicity, the following list describes only some substrings matched by the regular expressions for each technique:
-
-	\begin{itemize}
-		\item automated testing: testcase, test case, unit test, automated test, test result, run the test, test run;
-		\item source code inspection: inspecting, inspection, look at the code, looking at the source;
-		\item ad hoc testing: verified by running, verified by [word ending in ``ing''];
-		\item automated static analysis: findbugs, checkstyle, static code, static analys;
-		\item formal testing: formal test, formal verif.
-	\end{itemize}
-
-	% The classification script is smart enough to ignore negative assertions, such as ``did not run the test''.
-
-	During the informal analysis, it was observed that ad hoc testing is reported in many distinct ways that are difficult to capture by means of regular expressions.
-Hence, its regular expression is expected to match only a few comments.
-
-	% The regular expression for ad hoc testing is the most restrictive of them, because ad hoc testing was found to be reported in many distinct ways (when it is reported).
-%We expect to find fewer examples of ad hoc testing than it should be found.
-
-\end{subsection}
-
 \begin{subsection}{Analysis: When Are Bugs Verified?}
@@
 It is expected that, if the discovered set of developers is actually a QA team, they should be responsible for the majority of the verifications.
+
+\end{subsection}
+
+\begin{subsection}{Analysis: How Are Bugs Verified?}
+
+	In order to discover the verification techniques used by the subprojects, we have looked at the comments written by developers when they mark a bug as \VERIFIED\ (meaning that the fix was accepted) or \REOPENED\ (meaning that the fix was rejected).
+First, an informal analysis was performed, by reading a sample of such comments, looking for textual patterns that indicate that a particular technique was used.
+Then, based on the informal analysis and on previous knowledge, a regular expression was built for each technique, to match comments in which there is evidence that a particular technique was used.
+The regular expressions were refined by looking at the results matched by them and modifying them to avoid false positives.
+
+	The complete regular expressions are available in the experimental package.
+For the sake of simplicity, the following list describes only some substrings matched by the regular expressions for each technique:
+
+	\begin{itemize}
+		\item automated testing: testcase, test case, unit test, automated test, test result, run the test, test run;
+		\item source code inspection: inspecting, inspection, look at the code, looking at the source;
+		\item ad hoc testing: verified by running, verified by [word ending in ``ing''];
+		\item automated static analysis: findbugs, checkstyle, static code, static analys;
+		\item formal testing: formal test, formal verif.
+	\end{itemize}
+
+	% The classification script is smart enough to ignore negative assertions, such as ``did not run the test''.
+
+	During the informal analysis, it was observed that ad hoc testing is reported in many distinct ways that are difficult to capture by means of regular expressions.
+Hence, its regular expression is expected to match only a few comments.
+
+	% The regular expression for ad hoc testing is the most restrictive of them, because ad hoc testing was found to be reported in many distinct ways (when it is reported).
+%We expect to find fewer examples of ad hoc testing than it should be found.
 
@@
 % \end{subsection}
-
-\begin{subsection}{How Are Bugs Verified?}
-
-  % To identify the verification techniques used by the projects, we looked at messages written by developers when they mark a bug as VERIFIED (meaning that the bug fix is appropriate) or REOPENED (meaning that some problem was found in the proposed fix).
-First, we have read some messages, non-systematically, to get some idea of what is said in theses messages and how it is said.
-
-	Table \ref{tab:how} shows, for each subproject, the number of comments associated with a bug being marked as \VERIFIED\ or \REOPENED\ that refers to a particular verification technique.
-No references to formal verification or static analysis were found; therefore, these techniques are not shown on the table.
-
-	\begin{table}
-		\begin{center}
-		\caption{Verification techniques for all four subprojects.} \label{tab:how}
-		\begin{tabular}{l|r|r|r}
-			% \toprule
-			\textbf{\centering Project} & 
-			\textbf{\centering Testing} & 
-			\textbf{\centering Inspection} & 
-			\textbf{\centering Ad Hoc} \\
-			\hline
-			Eclipse/Platform        & 260 (1.1\%) & 524 (2.2\%) & 74 (0.3\%) \\
-			Eclipse/EMF             & 27 (0.7\%) & 2 (0.1\%) & 0 (0.0\%) \\
-			NetBeans/Platform       & 94 (0.6\%) & 5 (0.0\%) & 0 (0.0\%) \\
-			NetBeans/VersionControl & 4 (0.1\%) & 1 (0.0\%) & 0 (0.0\%) 
-			% \bottomrule
-		\end{tabular}
-		\end{center}
-	\end{table}
-
-	In Eclipse/Platform, less than 4\% of the comments referenced the verification technique that was used; for the other projects, less than 1\% of the comments contained such references.
-Such low proportion reveals how difficult it is to infer the verification technique that was applied using only data from bug repositories.
-This is probably tacit knowledge that can only be extracted by asking the developers themselves.
-
-	In all subprojects, comments suggest the use of automated testing and code inspection in the verification process, the former technique being more frequently referenced (except in Eclipse/Platform, in which code inspection is cited more often).
-Evidences of ad hoc testing were found only in Eclipse/Platform, because verification comments tend to be more structured in this project, but we believe that ad hoc testing is much more common, given its unstructured nature.
-
-	While sorting through comments, we have found interesting information regarding the software verification process.
-Most verification comments do not provide any clue about the verification technique used.
-Most often, developers just state that the bug fix was verified, sometimes informing the build used to verify the fix.
-
-	In Eclipse/Platform, the developer who fixes a bug often asks someone else to verify the fix, by asking ``please verify, [developer name]''.
-If the bug is reopened, then the fixer and the verifier exchange roles.
-This behavior illustrates a structured bug fixing/verification process.
-
-	In Eclipse/EMF, we found that marking a bug as \VERIFIED\ does not mean that the bug fix was verified.
-Instead, it means that the fix was made available in a build of the software that is published in the subproject's website.
-This reinterpretation of the Bugzilla workflow was decided in a meeting\footnote{\url{http://wiki.eclipse.org/Modeling_PMC_Meeting,_2007-10-16}}.
-
-  % In Eclipse/Platform, though, a significant number of verification messages include the technique used, e.g., ``verified by code inspection'', or ``verified with test case''.
-Based on such observations, we have built a search term for inspections and another one for unit tests.
-Then, we have classified the technique described in each verification message using the search term.
-
-	%   In Eclipse/Platform, there are 13,483 verifications, each one with a message.
-By the search criteria, 494 were identified as code inspections, 118 were identified as tests and 74 were identified as ad hoc.
-	% 
-	% Evidence of both code inspection and automated testing was found in all projects, although the evidence was weaker.
-
-	% \textbf{Observations}.
-	%   We also noted that, in Eclipse/Platform, often the fixer requests someone else to verify the fix, by asking ``please verify, [developer name]''.
-If the bug is reopened, then the fixer and the verifier exchange roles.
-That is considered good practice.
-	% 
-	%   It is difficult to identify ad hoc cases, because they are so pervasive that they are not reported, and also they are reported in many different ways.
-
-	% Eclipse/EMF: VERIFIED = in a build available in the website.
-
-	%We analyzed cases in which the bug is fixed, the fixer asks someone else to verify the fix, and then the bug is reopened.
-
-% \begin{itemize}
-% 	\item By creating or running unit tests.
-% 	\item By creating throw-away or configuration code that forces the occurrence of the error.
-% 	\item By running the software in a determined way (ad hoc). Often indicated by verbs.
-% 	\item By inspecting code.
-% \end{itemize}
-
-	% Conclusion: it's hard to determine the technique used by just looking at the data. Eclipse/Platform documents it better, sometimes.
-
-\end{subsection}
 
@@
 
+\begin{subsection}{How Are Bugs Verified?}
+
+  % To identify the verification techniques used by the projects, we looked at messages written by developers when they mark a bug as VERIFIED (meaning that the bug fix is appropriate) or REOPENED (meaning that some problem was found in the proposed fix).
+First, we have read some messages, non-systematically, to get some idea of what is said in theses messages and how it is said.
+
+	Table \ref{tab:how} shows, for each subproject, the number of comments associated with a bug being marked as \VERIFIED\ or \REOPENED\ that refers to a particular verification technique.
+No references to formal verification or static analysis were found; therefore, these techniques are not shown on the table.
+
+	\begin{table}
+		\begin{center}
+		\caption{Verification techniques for all four subprojects.} \label{tab:how}
+		\begin{tabular}{l|r|r|r}
+			% \toprule
+			\textbf{\centering Project} & 
+			\textbf{\centering Testing} & 
+			\textbf{\centering Inspection} & 
+			\textbf{\centering Ad Hoc} \\
+			\hline
+			Eclipse/Platform        & 260 (1.1\%) & 524 (2.2\%) & 74 (0.3\%) \\
+			Eclipse/EMF             & 27 (0.7\%) & 2 (0.1\%) & 0 (0.0\%) \\
+			NetBeans/Platform       & 94 (0.6\%) & 5 (0.0\%) & 0 (0.0\%) \\
+			NetBeans/VersionControl & 4 (0.1\%) & 1 (0.0\%) & 0 (0.0\%) 
+			% \bottomrule
+		\end{tabular}
+		\end{center}
+	\end{table}
+
+	In Eclipse/Platform, less than 4\% of the comments referenced the verification technique that was used; for the other projects, less than 1\% of the comments contained such references.
+Such low proportion reveals how difficult it is to infer the verification technique that was applied using only data from bug repositories.
+This is probably tacit knowledge that can only be extracted by asking the developers themselves.
+
+	In all subprojects, comments suggest the use of automated testing and code inspection in the verification process, the former technique being more frequently referenced (except in Eclipse/Platform, in which code inspection is cited more often).
+Evidences of ad hoc testing were found only in Eclipse/Platform, because verification comments tend to be more structured in this project, but we believe that ad hoc testing is much more common, given its unstructured nature.
+
+	While sorting through comments, we have found interesting information regarding the software verification process.
+Most verification comments do not provide any clue about the verification technique used.
+Most often, developers just state that the bug fix was verified, sometimes informing the build used to verify the fix.
+
+	In Eclipse/Platform, the developer who fixes a bug often asks someone else to verify the fix, by asking ``please verify, [developer name]''.
+If the bug is reopened, then the fixer and the verifier exchange roles.
+This behavior illustrates a structured bug fixing/verification process.
+
+	In Eclipse/EMF, we found that marking a bug as \VERIFIED\ does not mean that the bug fix was verified.
+Instead, it means that the fix was made available in a build of the software that is published in the subproject's website.
+This reinterpretation of the Bugzilla workflow was decided in a meeting\footnote{\url{http://wiki.eclipse.org/Modeling_PMC_Meeting,_2007-10-16}}.
+
+  % In Eclipse/Platform, though, a significant number of verification messages include the technique used, e.g., ``verified by code inspection'', or ``verified with test case''.
+Based on such observations, we have built a search term for inspections and another one for unit tests.
+Then, we have classified the technique described in each verification message using the search term.
+
+	%   In Eclipse/Platform, there are 13,483 verifications, each one with a message.
+By the search criteria, 494 were identified as code inspections, 118 were identified as tests and 74 were identified as ad hoc.
+	% 
+	% Evidence of both code inspection and automated testing was found in all projects, although the evidence was weaker.
+
+	% \textbf{Observations}.
+	%   We also noted that, in Eclipse/Platform, often the fixer requests someone else to verify the fix, by asking ``please verify, [developer name]''.
+If the bug is reopened, then the fixer and the verifier exchange roles.
+That is considered good practice.
+	% 
+	%   It is difficult to identify ad hoc cases, because they are so pervasive that they are not reported, and also they are reported in many different ways.
+
+	% Eclipse/EMF: VERIFIED = in a build available in the website.
+
+	%We analyzed cases in which the bug is fixed, the fixer asks someone else to verify the fix, and then the bug is reopened.
+
+% \begin{itemize}
+% 	\item By creating or running unit tests.
+% 	\item By creating throw-away or configuration code that forces the occurrence of the error.
+% 	\item By running the software in a determined way (ad hoc). Often indicated by verbs.
+% 	\item By inspecting code.
+% \end{itemize}
+
+	% Conclusion: it's hard to determine the technique used by just looking at the data. Eclipse/Platform documents it better, sometimes.
+
+\end{subsection}
+
 \end{section}

### 2012-03-28 contents.tex
Updating QA values after removing mass verifications.

@@
 			\textbf{\centering QA team size} & 
-			\MyHead{2.5cm}{\textbf{\% of verifications by QA team}} \\
+			\MyHead{1.5cm}{\textbf{\% of verifications by QA team}} \\
 			\hline
-			Eclipse/Platform        & 4 (2.4\%)  &  1.1\% \\
-			Eclipse/EMF             & 0 (0.0\%)  &  0.0\% \\
-			NetBeans/Platform       & 5 (21.7\%) & 93.3\% \\
-			NetBeans/VersionControl & 5 (20.8\%) & 93.2\% \\
+			Eclipse & & \\
+			\quad Platform        & 4 (2.4\%)  &  1.1\% \\
+			\quad EMF             & 0 (0.0\%)  &  0.0\% \\
+			NetBeans & & \\
+			\quad Platform       & 25 (18.8\%) & 80.1\% \\
+			\quad VersionControl & 5 (20.8\%) & 93.2\% \\
 			% \bottomrule
@@
 
-	% We counted the number of times each developer verified or fixed a bug.
-We consider that, if a developer is primarily concerned with quality assurance, then s/he verifies much more bugs that s/he fixes.
-	% 
-	% We defined ratio as ...
-	% 
-	% The verified/fixed ratio for the four projects is shown on Table XXX.
-
-% \begin{verbatim}
-% *** eclipse-platform 
-% Verifiers: 4 (2.4%)
-%   1.1% of verifications
-% Fixers: 30 (17.9%)
-% Total developers: 168 
-% 
-% *** eclipse-emf 
-% Verifiers: 0 (0.0%)
-%   0.0% of verifications
-% Fixers: 4 (30.8%)
-% Total developers: 13 
-% 
-% *** netbeans-platform 
-% Verifiers: 5 (21.7%)
-%   93.3% of verifications
-% Fixers: 6 (26.1%)
-% Total developers: 23 
-% 
-% *** netbeans-versioncontrol 
-% Verifiers: 5 (20.8%)
-%   93.2% of verifications
-% Fixers: 6 (25.0%)
-% Total developers: 24 
-% \end{verbatim}
-
-	In both NetBeans subprojects it is possible to infer the existence of a QA team, composed by approximately 20\% of the developers, that performs about 90\% of all verifications.
+	In both NetBeans subprojects it is possible to infer the existence of a QA team, composed by approximately 20\% of the developers, performing at least 80\% of all verifications.
 In the Eclipse subprojects, there is no evidence of a dedicated QA team.

### 2012-03-28 contents.tex
Further explanation on QA.

@@
 	Table \ref{tab:qa} presents, for each subproject, the number of developers attributed to a QA team (i.e., developers who perform verifications at least 10 times more often than contribute bug fixes), and the proportion of bug verifications they account for.
+Mass verifications and developers who have not contributed with fixes and verifications are discarded from the analysis.
+
 	\newcommand\MyHead[2]{%
@@
 In Eclipse/Platform, 2.4\% of the developers focus on verification tasks, contributing only to 1.1\% of the verifications.
+
+	Although the reported results are based on an arbitrary threshold (10) for the ratio between verifications and fixes, other values (2 and 5) were also used, leading to similar results.
 

### 2012-03-28 contents.tex
Writing about the removal of mass verifications.

@@
 
+	In the next two analyses (who and how), mass verifications were discarded.
+A verification was considered to be part of a mass verification if the developer who performed it also performed at least other 49 verifications in the same day.
+Although further research is needed to evaluate such criterion, it was able to identify the most obvious mass verification cases.
+
+	After applying the criterion to identify mass verifications, 362 (2.4\%) verifications were discarded from Eclipse/Platform, 2348 (72.3\%) verifications were discarded from Eclipse/EMF, and 5336 (39.3\%) verifications were discarded from NetBeans/Platform.
+Mass verifications were not identified in NetBeans/VersionControl.
+
 	% 
@@
 	Table \ref{tab:qa} presents, for each subproject, the number of developers attributed to a QA team (i.e., developers who perform verifications at least 10 times more often than contribute bug fixes), and the proportion of bug verifications they account for.
-Mass verifications and developers who have not contributed with fixes and verifications are discarded from the analysis.
+Mass verifications and developers who have not contributed with fixes and verifications were discarded from the analysis.
 

### 2012-03-28 contents.tex
Adding distribution of ratios (who).

@@
 		\caption{Discovered QA team for all four subprojects.} \label{tab:qa}
-		\begin{tabular}{l|r|r}
+		\begin{tabular}{l|r|r|c}
 			% \toprule
 			\textbf{\centering Project} & 
-			\textbf{\centering QA team size} & 
-			\MyHead{1.5cm}{\textbf{\% of verifications by QA team}} \\
+			\MyHead{1.2cm}{\textbf{\centering QA team size}} & 
+			\MyHead{1.6cm}{\textbf{\% of verifications by QA team}} &
+			\MyHead{2.7cm}{\textbf{Distribution of ratio (bugs verified / bugs fixed) per developer}} \\
 			\hline
-			Eclipse & & \\
-			\quad Platform        & 4 (2.4\%)  &  1.1\% \\
-			\quad EMF             & 0 (0.0\%)  &  0.0\% \\
-			NetBeans & & \\
-			\quad Platform       & 25 (18.8\%) & 80.1\% \\
-			\quad VersionControl & 5 (20.8\%) & 93.2\% \\
+			\textbf{Eclipse} & & & 
+			   \multirow{6}{*}{\includegraphics[scale=1]{ratio}}\\
+			\quad Platform        & 4 (2.4\%)  &  1.1\% &  \\
+			\quad EMF             & 0 (0.0\%)  &  0.0\% & \\
+			\textbf{NetBeans} & & \\
+			\quad Platform       & 25 (18.8\%) & 80.1\% & \\
+			\quad V.Control & 5 (20.8\%) & 93.2\% & \\
 			% \bottomrule
@@
 	Although the reported results are based on an arbitrary threshold (10) for the ratio between verifications and fixes, other values (2 and 5) were also used, leading to similar results.
+The rightmost column of Table \ref{tab:qa} shows the distribution of the ratio, in log scale, where it can be seen that ratios above 10 are uncommon in all projects.
 

### 2012-03-28 contents.tex
Reducing some text and rewriting some parts of the how results.

@@
 
-Bug tracking systems allow users and developers of a software project to manage a list of bugs for the project, along with information such as steps to reproduce the bug, its severity and the operating system used.
-Developers choose bugs to fix and can report on the progress of the bug fixing activities, ask for clarification, discuss causes for the bug etc.
+Bug tracking systems allow users and developers of a software project to manage a list of bugs for the project, along with information such as steps to reproduce the bug and the operating system used.
+Developers choose bugs to fix and report on the progress of the bug fixing activities, ask for clarification, discuss causes for the bug etc.
 
-In this research, we focus on Bugzilla, an open source bug tracking system used by notable software projects such as Eclipse, Mozilla, Linux Kernel, NetBeans, Apache, and companies such as NASA and Facebook\footnote{Complete list available at http://www.bugzilla.org/installation-list/.}.
+In this research, we focus on Bugzilla, an open source bug tracking system used by software projects such as Eclipse, Mozilla, Linux Kernel, NetBeans, Apache, and companies such as NASA and Facebook\footnote{Complete list available at http://www.bugzilla.org/installation-list/.}.
 The general concepts from Bugzilla should apply to most other bug tracking systems.
@@
 
-In the simplest cases, a bug is created and receive the status \UNCONFIRMED\ (if it was created by a regular user) or \NEW\ (if it was created by a developer).
+In simple cases, a bug is created and receive the status \UNCONFIRMED\ (when created by a regular user) or \NEW\ (when created by a developer).
 Next, it is \ASSIGNED\ to a developer, and then it is \RESOLVED, possibly by fixing it with a patch on the source code.
@@
 Bugzilla documentation states that, when a bug is \VERIFIED, it means that ``QA [quality assurance team] has looked at the bug and the resolution and agrees that the appropriate resolution has been taken''\footnote{\url{https://landfill.bugzilla.org/bugzilla-3.6-branch/page.cgi?id=fields.html}}.
-It does not specify how developers should look at the resolution (e.g., by looking at the source code, or by running a patched version of the software).
+It does not specify how developers should look at the resolution (e.g., by looking at the code, or by running the patched software).
 
@@
 	\begin{enumerate}
-		\item \textbf{Data extraction}: we have obtained publicly available raw data from the Bugzilla repositories of two popular integrated development environments, Eclipse and NetBeans.
+		\item \textbf{Data extraction}: we have obtained publicly available raw data from the Bugzilla repositories of two integrated development environments, Eclipse and NetBeans.
 
@@
 
-		\item \textbf{Data analysis}: for each research question, a distinct analysis was required. The analyses will be further described.
+		\item \textbf{Data analysis}: for each research question, a distinct analysis was required, as will be further described.
 	\end{enumerate}
@@
 
-	Table \ref{tab:how} shows, for each subproject, the number of comments associated with a bug being marked as \VERIFIED\ or \REOPENED\ that refers to a particular verification technique.
-No references to formal verification or static analysis were found; therefore, these techniques are not shown on the table.
+	Table \ref{tab:how} shows, for each subproject, the number of comments associated with a bug being marked as \VERIFIED\ or \REOPENED\ that refers to a particular verification technique, as matched by the respective regular expressions.
+Comments associated with mass verifications were discarded.
+
+	% No references to formal verification were found.
+	% 
+	% No references to formal verification or static analysis were found; therefore, these techniques are not shown on the table.
 
@@
 			\hline
-			Eclipse/Platform        & 260 (1.1\%) & 524 (2.2\%) & 74 (0.3\%) \\
-			Eclipse/EMF             & 27 (0.7\%) & 2 (0.1\%) & 0 (0.0\%) \\
-			NetBeans/Platform       & 94 (0.6\%) & 5 (0.0\%) & 0 (0.0\%) \\
-			NetBeans/VersionControl & 4 (0.1\%) & 1 (0.0\%) & 0 (0.0\%) 
+			Eclipse/Platform        & 250 (1.1\%) & 511 (2.2\%) & 67 (0.3\%) \\
+			Eclipse/EMF             & 21 (1.6\%) & 1 (0.1\%) & 0 \hspace{0.8cm} \\
+			NetBeans/Platform       & 88 (0.8\%) & 5 (0.0\%) & 0 \hspace{0.8cm} \\
+			NetBeans/VersionControl & 4 (0.1\%) & 1 (0.0\%) & 0 \hspace{0.8cm}
 			% \bottomrule
@@
 
-	In Eclipse/Platform, less than 4\% of the comments referenced the verification technique that was used; for the other projects, less than 1\% of the comments contained such references.
+	No references to formal verification were found.
+Regarding static analysis, only 4 references were found; upon inspection, it was found that only one reference (bug report #15242 for NetBeans/Platform) implies the use of a static analysis tool in the verification process.
+
+	In all subprojects, comments suggest the use of automated testing and code inspection in the verification process, the former technique being more frequently referenced (except in Eclipse/Platform, in which code inspection is cited more often).
+Evidences of ad hoc testing were found only in Eclipse/Platform, probably due to limitations in the regular expressions.
+
+	% Few comments
+	% 
+	The regular expressions were able to identify the verification technique only in 3.6\% of the comments at best (Eclipse/Platform).
+Such low proportion can be explained partly by limitations of the regular expression method (which can be addressed in future work) and partly by lack of information in the comments themselves.
+	% In future work, we expect to improve the results by using more sophisticated information retrieval techniques.
+%We suspect, however, that the majority of the comments do not provide relevant information about the verification technique that was used.
+
+	By sorting through comments, we have found that many of them do not provide any clue about the verification technique used.
+Most often, developers just state that the bug fix was verified, sometimes informing the build used to verify the fix.
+
+	% In Eclipse/Platform, less than 4\% of the comments referenced the verification technique that was used; for the other projects, the proportion was even lower.
 Such low proportion reveals how difficult it is to infer the verification technique that was applied using only data from bug repositories.
@@
 
-	In all subprojects, comments suggest the use of automated testing and code inspection in the verification process, the former technique being more frequently referenced (except in Eclipse/Platform, in which code inspection is cited more often).
-Evidences of ad hoc testing were found only in Eclipse/Platform, because verification comments tend to be more structured in this project, but we believe that ad hoc testing is much more common, given its unstructured nature.
+	% Four references to static analysis tools were found in the comments.
+Upon further investigation, it was discovered that only in one case (bug report #15242 for NetBeans/Platform) such tool was said to be used for verification purposes.
 
-	While sorting through comments, we have found interesting information regarding the software verification process.
-Most verification comments do not provide any clue about the verification technique used.
+  % While sorting through comments, we have found interesting information regarding the software verification process.
+Many verification comments do not provide any clue about the verification technique used.
 Most often, developers just state that the bug fix was verified, sometimes informing the build used to verify the fix.
 
-	In Eclipse/Platform, the developer who fixes a bug often asks someone else to verify the fix, by asking ``please verify, [developer name]''.
+	In Eclipse/Platform, comments show that the developer who fixes a bug often asks someone else to verify the fix, by asking ``please verify, [developer name]''.
 If the bug is reopened, then the fixer and the verifier exchange roles.
@@
 	In Eclipse/EMF, we found that marking a bug as \VERIFIED\ does not mean that the bug fix was verified.
-Instead, it means that the fix was made available in a build of the software that is published in the subproject's website.
-This reinterpretation of the Bugzilla workflow was decided in a meeting\footnote{\url{http://wiki.eclipse.org/Modeling_PMC_Meeting,_2007-10-16}}.
+Instead, it means that the fix was made available in a build of the software that is published in the subproject's website\footnote{See \url{http://wiki.eclipse.org/Modeling_PMC_Meeting,_2007-10-16}}.
 

### 2012-03-28 contents.tex
Simplifying description of how method.

@@
 
-	In order to discover the verification techniques used by the subprojects, we have looked at the comments written by developers when they mark a bug as \VERIFIED\ (meaning that the fix was accepted) or \REOPENED\ (meaning that the fix was rejected).
-First, an informal analysis was performed, by reading a sample of such comments, looking for textual patterns that indicate that a particular technique was used.
+	In order to discover the verification techniques used by the subprojects, we have selected the comments written by developers when they mark a bug as \VERIFIED\ (meaning that the fix was accepted) or \REOPENED\ (meaning that the fix was rejected).
+
+	The comments were matched against five regular expressions, each corresponding to one of the following verification techniques: automated testing, source code inspection, ad hoc testing, automated static analysis, and formal testing.
+The complete regular expressions are available in the experimental package.
+
+	It should be noted that regular expressions may not be a reliable alternative to the problem of identifying verification techniques in comments.
+In future research, more advanced information retrieval techniques should be explored.
+Nevertheless, regular expressions should enable an initial study and help unveil insights about verification techniques in bug reports.
+
+	% identify reliably the techniques prese
+
+	% A regular expression was constructed to match terms related to each verification technique.
+	% 
+	% First, an informal analysis was performed, by reading a sample of such comments, looking for textual patterns that indicate that a particular technique was used.
 Then, based on the informal analysis and on previous knowledge, a regular expression was built for each technique, to match comments in which there is evidence that a particular technique was used.
 The regular expressions were refined by looking at the results matched by them and modifying them to avoid false positives.
-
-	The complete regular expressions are available in the experimental package.
+	% 
+	% The complete regular expressions are available in the experimental package.
 For the sake of simplicity, the following list describes only some substrings matched by the regular expressions for each technique:
-
-	\begin{itemize}
-		\item automated testing: testcase, test case, unit test, automated test, test result, run the test, test run;
-		\item source code inspection: inspecting, inspection, look at the code, looking at the source;
-		\item ad hoc testing: verified by running, verified by [word ending in ``ing''];
-		\item automated static analysis: findbugs, checkstyle, static code, static analys;
-		\item formal testing: formal test, formal verif.
-	\end{itemize}
+	% 	
+	% \begin{itemize}
+	% 	\item automated testing: testcase, test case, unit test, automated test, test result, run the test, test run;
+	% 	\item source code inspection: inspecting, inspection, look at the code, looking at the source;
+	% 	\item ad hoc testing: verified by running, verified by [word ending in ``ing''];
+	% 	\item automated static analysis: findbugs, checkstyle, static code, static analys;
+	% 	\item formal testing: formal test, formal verif.
+	% \end{itemize}
 
@@
 
-	During the informal analysis, it was observed that ad hoc testing is reported in many distinct ways that are difficult to capture by means of regular expressions.
+	% During the informal analysis, it was observed that ad hoc testing is reported in many distinct ways that are difficult to capture by means of regular expressions.
 Hence, its regular expression is expected to match only a few comments.
@@
 
-	In all subprojects, comments suggest the use of automated testing and code inspection in the verification process, the former technique being more frequently referenced (except in Eclipse/Platform, in which code inspection is cited more often).
+	In all projects, comments suggest the use of automated testing and code inspection in the verification process, the former technique being more frequently referenced (except in Eclipse/Platform).
 Evidences of ad hoc testing were found only in Eclipse/Platform, probably due to limitations in the regular expressions.

### 2012-03-29 contents.tex
Changing bugzilla workflow figure to a more compact one.

@@
 However, mining bug repositories has its own risks.
-Previous research has identified problems of missing data (e.g., rationale, traceability links between reported bug fixes and source code changes), inaccurate data (e.g., misclassification of bugs) \cite{Aranda2009}, and biased data \cite{Bird2009}.
+Previous research has identified problems of missing data \cite{Arand2009} (e.g., rationale, traceability links between reported bug fixes and source code changes), inaccurate data (e.g., misclassification of bugs) \cite{Antoniol2008}, and biased data \cite{Bird2009}.
 %,Bird2009,Nguyen2010,Ayari2007}.
@@
 	\centering
-		\includegraphics[scale=0.4]{bugzilla.png}
-	\caption{Workflow for Bugzilla. Source: \url{http://www.bugzilla.org/docs/2.18/html/lifecycle.html}.}
+		\includegraphics[width=0.48\textwidth]{bugzilla}
+	\caption{Simplified workflow for Bugzilla. Adapted from \url{http://www.bugzilla.org/docs/2.18/html/lifecycle.html}.}
 	\label{fig:bugzilla}
@@
 			\MyHead{1.6cm}{\textbf{\% of verifications by QA team}} &
-			\MyHead{2.7cm}{\textbf{Distribution of ratio (bugs verified / bugs fixed) per developer}} \\
+			\MyHead{2.7cm}{\textbf{Distribution of ratio (bugs verified / bugs fixed) per developer}\vspace{2mm}} \\
 			\hline
@@
 	No references to formal verification were found.
-Regarding static analysis, only 4 references were found; upon inspection, it was found that only one reference (bug report #15242 for NetBeans/Platform) implies the use of a static analysis tool in the verification process.
+Regarding static analysis, only 4 references were found; upon inspection, it was found that only one reference (bug report 15242 for NetBeans/Platform) implies the use of a static analysis tool in the verification process.
 

### 2012-03-29 contents.tex
Adding threats to validity, changing when figure, searching for 'QA team' to confirm results about who.

@@
 However, mining bug repositories has its own risks.
-Previous research has identified problems of missing data \cite{Arand2009} (e.g., rationale, traceability links between reported bug fixes and source code changes), inaccurate data (e.g., misclassification of bugs) \cite{Antoniol2008}, and biased data \cite{Bird2009}.
+Previous research has identified problems of missing data \cite{Aranda2009} (e.g., rationale, traceability links between reported bug fixes and source code changes), inaccurate data (e.g., misclassification of bugs) \cite{Antoniol2008}, and biased data \cite{Bird2009}.
 %,Bird2009,Nguyen2010,Ayari2007}.
@@
 \centering
-\mbox{
-\subfigure{\includegraphics[width=0.25\textwidth]{milestones}}%\quad
-\subfigure{\includegraphics[width=0.25\textwidth]{cliffs} }}
-\caption{Number of verifications over time.} \label{fig:when}
+\includegraphics[width=0.48\textwidth]{when}
+% \mbox{
+% \subfigure{\includegraphics[width=0.25\textwidth]{milestones}}%\quad
+% \subfigure{\includegraphics[width=0.25\textwidth]{cliffs} }}
+\caption{Accumulated number of verifications over time.} \label{fig:when}
 \end{figure}
@@
 
-	Table \ref{tab:qa} presents, for each subproject, the number of developers attributed to a QA team (i.e., developers who perform verifications at least 10 times more often than contribute bug fixes), and the proportion of bug verifications they account for.
+	Table \ref{tab:qa} presents, for each subproject, the number of developers attributed to a QA team (i.e., developers with verifications per fix ratio greater than the threshold of 10), and the proportion of bug verifications they account for.
+Other threshold values (2 and 5) were also tried for the ratio, leading to similar results.
 Mass verifications and developers who have not contributed with fixes and verifications were discarded from the analysis.
@@
 In the Eclipse subprojects, there is no evidence of a dedicated QA team.
-In Eclipse/EMF no developer focus specifically on verification.
-In Eclipse/Platform, 2.4\% of the developers focus on verification tasks, contributing only to 1.1\% of the verifications.
+Further evidence was found by looking for the substring ``QA team'' in the comments associated with the status \VERIFIED\ and \REOPENED.
+Comments referring to a QA team were only found in NetBeans subprojects.
 
-	Although the reported results are based on an arbitrary threshold (10) for the ratio between verifications and fixes, other values (2 and 5) were also used, leading to similar results.
+	% In Eclipse/EMF no developer focus specifically on verification. In Eclipse/Platform, 2.4\% of the developers focus on verification tasks, contributing only to 1.1\% of the verifications.
+
+	% Although the reported results are based on an arbitrary threshold (10) for the ratio between verifications and fixes, other values (2 and 5) were also used, leading to similar results.
 The rightmost column of Table \ref{tab:qa} shows the distribution of the ratio, in log scale, where it can be seen that ratios above 10 are uncommon in all projects.
@@
 
+	The main threat to this study is related to construct validity, i.e., to which point the operational definitions of concepts such as QA team and verification phase reflect the theoretical concepts.
+To mitigate this problem, we have looked for additional information in the text of comments in the bug reports.
+Further evidence can be gathered by interviewing developers and mining source code repositories.
+
 	With the knowledge obtained from this exploratory research, we aim to improve and extend our previous work on the impact of independent verification on software quality.
@@
 
-	Researchers should be aware that information about verification techniques may not be common in bug repositories, and that reported verification does not always correspond to actual verification.
+	Researchers should be aware that information about verification techniques may not be common in bug repositories, and that reported verification does not always correspond to actual verification (e.g., in the case of mass verifications).
 Some exploration of the data is important to avoid such pitfalls.

### 2012-03-29 contents.tex
Overall revision.

@@
 However, mining bug repositories has its own risks.
-Previous research has identified problems of missing data \cite{Aranda2009} (e.g., rationale, traceability links between reported bug fixes and source code changes), inaccurate data (e.g., misclassification of bugs) \cite{Antoniol2008}, and biased data \cite{Bird2009}.
+Previous research has identified problems of missing data (e.g., rationale, traceability links between reported bug fixes and source code changes) \cite{Aranda2009}, inaccurate data (e.g., misclassification of bugs) \cite{Antoniol2008}, and biased data \cite{Bird2009}.
 %,Bird2009,Nguyen2010,Ayari2007}.
@@
 In future research, more advanced information retrieval techniques should be explored.
-Nevertheless, regular expressions should enable an initial study and help unveil insights about verification techniques in bug reports.
+Nevertheless, regular expressions enable an initial study and help unveil insights about verification techniques in bug reports.
 
@@
 
-	Each of the three analysis was performed on Eclipse/Platform, Eclipse/EMF, NetBeans/Platform, and NetBeans/VersionControl. The results are presented next.
+	Each of the three analyses was performed on Eclipse/Platform, Eclipse/EMF, NetBeans/Platform, and NetBeans/VersionControl. The results are presented next.
 
@@
 
-	In both NetBeans subprojects it is possible to infer the existence of a QA team, composed by approximately 20\% of the developers, performing at least 80\% of all verifications.
 	In the Eclipse subprojects, there is no evidence of a dedicated QA team.
+In both NetBeans subprojects, on the other hand, it is possible to infer the existence of a QA team, composed by approximately 20\% of the developers, performing at least 80\% of all verifications.
 Further evidence was found by looking for the substring ``QA team'' in the comments associated with the status \VERIFIED\ and \REOPENED.
@@
 
-	By sorting through comments, we have found that many of them do not provide any clue about the verification technique used.
+	By reading comments, we have found that many of them mention any verification technique.
 Most often, developers just state that the bug fix was verified, sometimes informing the build used to verify the fix.
@@
 If the bug is reopened, then the fixer and the verifier exchange roles.
-This behavior illustrates a structured bug fixing/verification process.
+This behavior illustrates a structured bug fixing/verification process and supports the conclusion that there is no QA team in Eclipse/Platform.
 
@@
 
-\small{This work is supported by FAPESB (grant BOL0119/2010) and CNPq/INES (grant 573964/2008-4).}
+%\small{
+This work is supported by FAPESB under grant BOL0119/2010 and by CNPq/INES under grant 573964/2008-4.% er wer wer wer we rwe rwe rwe rwe rwe rw erw er wer wer wer wer wer wer wer wer  qwe qwe qwe qwe qwe qwe qwe qw eqwe  qwe qwe qwe qwe q qwe qwe qwe qw e
+%}

### 2012-03-29 contents.tex
Fixing problem with margins (reported by the Conference Publishing script). Camera ready copy.

@@
 	\centering
+	\vspace{2mm}
 		\includegraphics[width=0.48\textwidth]{bugzilla}
@@
 \centering
+\vspace{2mm}
 \includegraphics[width=0.48\textwidth]{when}
@@
 	\begin{table}
+		\vspace{2mm}
 		\begin{center}
@@
 		\begin{center}
+		\vspace{2mm}
 		\caption{Verification techniques for all four subprojects.} \label{tab:how}
@@
 
-%\small{
+\small{
 This work is supported by FAPESB under grant BOL0119/2010 and by CNPq/INES under grant 573964/2008-4.% er wer wer wer we rwe rwe rwe rwe rwe rw erw er wer wer wer wer wer wer wer wer  qwe qwe qwe qwe qwe qwe qwe qw eqwe  qwe qwe qwe qwe q qwe qwe qwe qw e
-%}
+}
